---
title: "DEPLOYMENT PERFORMANCE REVIEW OF THE 2024 NORTH PACIFIC OBSERVER PROGRAM"
always_allow_html: true
output:
  word_document:
    reference_docx: docs/word-styles-reference-01.docx
    pandoc_args:
      '--lua-filter=pagebreaks.lua'
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
# Once report has been run and 4_AR_report.Rdata has been created,
# skip running the code below and load all packages and objects here:
# load("4_AR_report.Rdata")
# source("3_helper.R")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load packages and functions 
source("3_helper.R")
# Load spatiotemporal functions 
source("functions/spatiotemp_functions.R")
```

```{r, include=FALSE, eval=FALSE}
#' Download data. Run this line manually during development, but it won't run during knitting.
gdrive_download("2_AR_data.Rdata", gdrive_set_dribble("Projects/AnnRpt-Deployment-Chapter"))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#' Load data and set parameters
load("2_AR_data.RData")

# Avoid showing code
knitr::opts_chunk$set(echo = FALSE)

# Avoid scientific notation
options(scipen = 10000)

# Avoid dplyr summarise warnings
options(dplyr.summarise.inform = FALSE)

# Set table output default settings
set_flextable_defaults(font.family = "Times New Roman", font.size = 11)

# Set order of strata
strata.order <- c("Full", "OB FIXED BSAI", "OB FIXED GOA", "OB TRW BSAI", "OB TRW GOA", "EM FIXED BSAI", "EM FIXED GOA", "EM TRW GOA (EFP)", "Zero coverage", "Total")

# Set the simulation parameters for the spatiotemporal analyses
sim_iter <- 1e4      # 1e3 iterations is good for testing quickly, 1e4 is better for the final run
seed <- 12345

# Is this the annual report or the tech memo?
# This will control whether certain text chunks are included and formatting of table/figure captions.
annual.report <- TRUE
tech.memo     <- FALSE

if(annual.report){
  fig_nums <- captioner(prefix = "Figure 3-", auto_space = FALSE)  # In Annual report, deployment analyses make up Chapter 3. 
  tbl_nums <- captioner(prefix = "Table 3-", auto_space = FALSE)
}
```

```{r set-options, message=FALSE, warning=FALSE}
#Total deployment strata:
# 2 full coverage (FULL, EM TRW EFP[BSAI])
# 4 partial coverage observed (OB FIXED BSAI and GOA, OB TRW BSAI and GOA)
# 3 partial coverage EM (EM FIXED BSAI and GOA, EM TRW EFP[GOA]) + 
# 0 zero coverage EM research (Zero EM Research) +
# 1 zero coverage (ZERO)
total.strat <- nrow(unique(work.data[, .(COVERAGE_TYPE, STRATA)]))
```

`r if(tech.memo){'#Abstract'}`  
`r if(tech.memo){'This report contains the analyses and findings of the Alaska Fisheries Science Center’s Fisheries Monitoring and Analysis Division’s Fishery Monitoring Science Committee (FMSC) on the efficiency and effectiveness of observer deployment following the 2024 Annual Deployment Plan (ADP). Responses to comments by the North Pacific Fishery Management Council (Council) from the 2022 version of this report, and recommendations to improve data quality and guide the 2024 Annual Deployment Plan are also included. In 2024, there were 10 strata to evaluate: two full coverage strata, five (*THREE IF TENDER DROPPED*) partial coverage observer strata defined by gear and tender designation (*ARE WE COVERING TENDERS STILL?*), three partial coverage Electronic Monitoring (EM) strata defined by gear designation, one zero coverage EM research stratum, and one zero coverage stratum. Observers were deployed under trip-selection on *UPDATE:*161 full coverage vessels that fished for *UPDATE:*3,343 trips and *UPDATE:*584 unique partial coverage vessel and stratum combinations (vessels can fish in more than one stratum) that fished *UPDATE:*5,016 trips total. This was the *UPDATE:*second year in which data from the *EM HAL* stratum was used in catch accounting, and the *UPDATE:*first year in which data from the *EM POT* stratum was used in catch accounting. *UPDATE:*The *EM HAL* stratum was the third largest stratum by trip count in 2019 with 916 total trips fished. EM systems were deployed onto 21 pot vessels that fished for 165 trips. *UPDATE:*Research EM systems were deployed onto four vessels that fished for 29 trips. *UPDATE:*A total of 393 vessels fished 2,005 trips in the *Zero Coverage* stratum, accounting for 28% of all trips that occurred within the partial coverage category. Realized coverage rates met expectations in *UPDATE:*eight of ten strata in 2024. The *Full* stratum was observed at a rate of 99.9%, falling below the expected rate of 100%. *UPDATE:*The partial coverage *POT – Tender* stratum was observed at a rate of 29.5%, with a 95% confidence interval that fell above the expected coverage rate of 16.1%. *UPDATE:*However, deployment in 2019 generally proceeded as planned, with few missed expectations or signs of potential bias. The FMSC recommends that future ADPs fully integrate EM and observer deployment into one fishery monitoring program. Although EM data are now being used in catch accounting, the funding and selection of EM vessels has so far occurred separately from the processes used to optimize observer deployment in each ADP. Considering EM as a more integral part of those optimization processes will better enable analysts to avoid data gaps and take advantage of the respective abilities of each monitoring method (observers and EM). The FMSC also continues to recommend that NMFS link the ODDS and eLandings database such that fishing trips can be uniquely identified to support the analyses presented to the Council. Such a linkage will better enable analysts to discern the contributing factors behind instances in which intended deployment is inconsistent with realized deployment.'}`  

# Introduction  
`r if(tech.memo){'##Background of the North Pacific Groundfish and Halibut Observer Program'}`  
`r if(tech.memo){'Fisheries observers and electronic monitoring (EM) systems collect independent information that is used to determine the effects of fishing on natural resources. The National Marine Fisheries Service (NMFS) uses its observer program in Alaska to enable the use of tools such as catch quotas to manage against the over- or under-harvest of fishes. Observers and EM are two verifiable methods for collecting fishery discard information used to estimate total catch. Observers (but not the EM systems currently used in the North Pacific) are able to record seabird and marine mammal interactions with fisheries as well. Observers also collect biological information such as length, sex, weight, ageing structures (e.g., otoliths, spines, scales, and vertebrae), and stomachs to support ecosystem studies and stock assessments.'}`

`r if(tech.memo){'The observer program in the North Pacific has a long history. Observers were first deployed onto fishing vessels in the Bering Sea in 1973 and into the remainder of the North Pacific in 1975 (Nelson et al. 1981, Wall et al. 1981). Fisheries in the North Pacific were initially prosecuted exclusively by foreign and later by "joint venture"'}` `r if(tech.memo){"operations where a developing domestic fleet of catcher vessels delivered to foreign-owned processing vessels. During the foreign and joint venture operations, foreign vessels carried fisheries observers at their expense, while domestic vessels were exempted from this observer coverage. As foreign vessels' rights to fish in the U.S. Exclusive Economic Zone (EEZ) were reduced over time and the domestic fishery grew, it became obvious to managers that observer coverage would be necessary for the emerging domestic fleet. At the onset of fully domestic fishery operations in 1990, the North Pacific Groundfish Observer Program was established as an interim observer program with rules governing observer coverage codified in regulations. This interim program would be extended four times over the next 20 years by the North Pacific Fishery Management Council (Council) — the last without a sunset date."}`

`r if(tech.memo){'The regulations established in 1990 required vessels 60-125 feet in length (overall) and all vessels fishing pot gear to carry observers at their own cost for 30% of their fishing days in a calendar quarter plus at least one trip in each fishery they participate in (termed the "30% fleet"), and vessels greater than 125 feet in length to carry an observer for 100% of their fishing days at their expense. Some vessels were not required to carry observers. These included vessels less than 60 feet, vessels fishing jig gear or vessels fishing with trawl gear that deliver unsorted codends to processing vessels (termed "catcher processors" or CPs if the vessel also has catching ability and "mothership" or M if the vessel does not) and vessels that fished for Pacific halibut (*Hippoglossus stenolepis*). For shoreside processors, the rules governing observer coverage were based on the estimated tonnage processed in a calendar month: plants that processed less than 500 metric tons (t) a month were exempted from coverage, those that processed between 500 t and 1,000 t a month were required to be observed for 30% of the calendar days fish, and those that processed more than 1,000 t a month were required to be observed for each day in the month.'}`

`r if(tech.memo){'Soon after the establishment of the domestic observer program, concerns over the ability and incentive for fishers to manipulate observer coverage in a way that might bias catch estimates and other analytic products prompted efforts by NMFS and the Council to provide a mechanism for NMFS to gain control over where and when observers were deployed (Faunce and Barbeaux 2011). From 1992 to 2008, several attempts to “restructure” the program were made. In 2010, the Council unanimously decided to move forward with the restructured observer program. In 2012, the Final Rule 77 FR 70062 was published to implement Amendment 86 to the Fishery Management Plan for Groundfish of the Bering Sea and Aleutian Islands (BSAI) Management Area and Amendment 76 to the Fishery Management Plan for Groundfish of the Gulf of Alaska (GOA). Amendments 86/76 added a funding and deployment system for observer coverage to the existing North Pacific Groundfish Observer Program and amended existing observer coverage requirements for vessels and processing plants. The “restructured” North Pacific Groundfish and Halibut Observer Program (hereafter termed “Observer Program”) began in 2013 with the randomization of deployments among trips and vessels. In 2018, the use of EM was added as an additional catch monitoring tool, with the understanding that some data elements collected by observers would not be collected using EM systems.'}`

`r if(tech.memo){'#The Annual Deployment Plan and Review'}`
`r if(tech.memo){"Analysis and evaluation of the data collected by observers is an ongoing process. The NMFS considers Council input in making decisions as to the amount of coverage (i.e., selection probabilities that are assigned to each partial-coverage category). These decisions are based on available funding, the cost of observer coverage, and anticipated effort. The restructure of the Observer Program established new annual reporting processes. Each June, the NMFS provides the Council with a comprehensive evaluation of past year's observer deployments, costs, sampling levels, and implementation issues as well as recommended changes for the coming year. The June deployment performance review aims to identify areas where improvements are needed to 1) collect the data necessary to manage the groundfish and halibut fisheries; 2) maintain the scientific goals of unbiased data collection; and 3) accomplish the most effective and efficient use of the funds collected through the observer fee. The annual deployment performance review is an opportunity to inform the Council and the public of how well various aspects of the program are working, and consequently lead to recommendations for improvement as appropriate. The NMFS also prepares the Observer Program Annual Deployment Plan (ADP) each fall. The ADP defines deployment strata and establishes selection rates given available budgets and anticipated fishing effort. A draft ADP is released by September of each year to allow review by the Council's Groundfish Plan Teams, as well as the Council and its Scientific and Statistical Committee (SSC). Based on input from its advisory bodies and the public, the Council may choose to clarify objectives and provide recommendations to NMFS for the ADP. Upon analysis of the Council recommendations, NMFS will make any necessary adjustments to finalize the ADP and release it to the public. The ADP is released to the public prior to the December Council meeting."}`    

`r if(tech.memo){'##Fishery Monitoring Science Committee'}`
`r if(tech.memo){'Each year the Alaska Fisheries Science Center’s (AFSC) Fisheries Monitoring and Analysis (FMA) Division establishes a committee to review the scientific elements of the North Pacific Observer Program. This committee, formerly referred to as the Observer Science Committee (OSC), was renamed in 2020 as the Fishery Monitoring Science Committee (FMSC), in order to reflect the addition of EM as a tool being used to monitor fisheries in the North Pacific. Similarly, we use the term ‘monitoring’ in this chapter when referencing fishing activity that has been monitored either by an observer or with EM.'}`  

The goal of the Observer Program is to achieve a random deployment of observers and electronic monitoring (EM) into Federal fisheries off Alaska to collect representative data used to estimate catch and bycatch, assess stock status, collect fishery-dependent biological information used in population and ecosystem modeling efforts, and make salmon bycatch stock-of-origin determinations, among other objectives. This chapter contains a review of the deployment of observers and EM in `r year` relative to the intended sampling plan and goals of the `r year` Annual Deployment Plan (ADP; NMFS `r year - 1`b). Consistent with its purpose, this chapter focuses on the randomization of observer and EM deployments into primary sampling units and how departures from a random sample affect data quality. This review identifies where possible biases exist and provides recommendations for further evaluation, including potential improvements to the observer deployment process that should be considered during the development of the `r year + 2` ADP.\
\
This review is performed by staff from the Fisheries Monitoring and Analysis/Analytical Services Program (FMA) of the Alaska Fisheries Science Center (AFSC) and the Sustainable Fisheries Division/Catch Analysis and Data Quality Branch of the Alaska Regional Office (AKRO). Catch and monitoring data from the `r year` calendar year as of 23 April `r year + 1` were used in analyses.\
\
In the past, the Fisheries Monitoring and Science Committee (FMSC) reviewed the analyses in this chapter and provided recommendations. However, for 2024 we... *ADD WHY WE DIDN’T DO THIS THIS YEAR, CIE REVIEW, ETC*\

## The Sampling Design of the Observer Program 
Since 2013, the Observer Program has used a stratified hierarchical sampling design with randomization at all levels (Cahalan and Faunce 2020). Stratification increases the efficiency of sampling by observers and helps address some logistical issues associated with deployment. By grouping similar fishing activities into strata and sampling those strata appropriately, sampling efficiency increases and the estimated variance decreases relative to unstratified sampling. Sampling strata are defined in the ADP and are designed such that each unit of deployment (e.g., trip) is assigned to only one stratum.\
\
Randomization helps ensure that the data collected from a sample will be representative of the entire fishing fleet (observed and monitored trips are equivalent to unobserved and unmonitored trips within a stratum). In each stratum, observers or EM are deployed randomly to either: (1) vessels for a predetermined period of time (termed 'vessel-selection') or (2) to individual fishing trips or shoreside deliveries of catch (termed 'trip-selection'). In both cases, this initial deployment to the fishery is the first level of the sampling hierarchy and defines the primary sampling unit (PSU; either vessel-periods or individual trips). The list of all PSUs in a stratum defines the sampling frame and should equate to the population of interest for that sampling stratum (e.g., all trips taken by trawl vessels fishing in the Alaska Exclusive Economic Zone). If the sampling frame does not contain all elements of the stratum, the resulting information may be biased. The magnitude and direction of the bias will depend on how different the fishing activities in the sample frame are from actual fishing activity.\
\
Although this chapter evaluates whether monitoring goals were met, we include a brief summary of the full sampling hierarchy here for context. For each observed trip, if all hauls cannot be sampled for logistical reasons, hauls are randomly selected to be sampled. This is the next level in the hierarchy; the secondary sampling units are defined as hauls within a trip. Randomization of haul selection is designed to allow observers time to record and transmit data, attend to other non-sampling responsibilities, and time to sleep and eat. Randomization of haul selection also gives EM video reviewers the ability to optimize the amount of video that can be reviewed from each trip. Haul selection is determined using the random sampling tables and random break tables provided by NMFS. For each haul, regardless of monitoring status, fishing location and effort (e.g., number of hooks) are recorded, while marine mammal and seabird interactions are primarily recorded on hauls randomly selected for monitoring. The ability of EM to capture marine mammal and seabird interactions is less than that of observers due to the fixed location in which EM equipment is installed.\
\
For the randomly selected hauls, a random sample of the catch is collected (observers) or selected for video review (EM) and data from those samples are used to determine the species composition and amount of discarded catch. These samples of catch within each haul are the third level of the sampling hierarchy. While observers are trained to collect multiple large samples of catch, the number and size of samples taken from each haul will depend on the vessel configuration, fishing operations, and diversity of catch. The size of EM samples is largely determined by the number of video reviewers available relative to the amount of video to be reviewed.\
\
At the fourth level of the sampling hierarchy, a predetermined number of individual fish of predetermined species are randomly selected from the species composition sample and measured. Lastly, at the fifth sampling level, a random selection of fish is used to collect otoliths, reproductive maturity assessments, stomach contents, genetic tissues, and other biological specimens. The number and species of fish selected for measurement and biological specimen collection is specified each year by the AFSC’s stock assessment scientists. Sampling rates for genetic tissue collection by observers (e.g., 1 in 10 Chinook salmon [*Oncorhynchus tshawytscha*] caught as bycatch) are set each year by the AFSC’s Auke Bay Laboratory. Sampling at the fourth and fifth levels of the sampling hierarchy does not occur with EM.

`r if(tech.memo){'#The'}` `r if(tech.memo){year}` `r if(tech.memo){'Annual Deployment Plan'}`    
`r if(tech.memo){'The following briefly summarizes the final'}` `r if(tech.memo){year}` `r if(tech.memo){'ADP (NMFS '}` `r if(tech.memo){year - 1}``r if(tech.memo){'). In general, all vessels that participate in cooperatives or act as catcher-processors or motherships are fully observed at the trip-level and constitute the full-coverage category of the fleet. In 2016, NMFS published new regulations to allow the owner of a trawl catcher vessel to annually request that NMFS place requesting vessels in the full coverage category for all directed fishing for groundfish using trawl gear in the Bering Sea and Aleutian Islands management area (BSAI) in the following calendar year. This regulated process has replaced an interim policy. For the '}` `r if(tech.memo){year}` `r if(tech.memo){'calendar year, NMFS received and approved requests and has placed'}` `r if(tech.memo){'32'}` `r if(tech.memo){'catcher vessels in the full coverage category for all directed fishing for groundfish using trawl gear in the BSAI management area (NMFS'}` `r if(tech.memo){year-1}``r if(tech.memo){'). The partial-coverage category includes vessels greater than or equal to 40 feet (ft) length overall (LOA) and not in the full coverage category. The following sampling strata comprised the partial coverage category in'}` `r if(tech.memo){year}` `r if(tech.memo){':'}`  

`r if(tech.memo){'1. Hook-and-line vessels (*OB HAL* stratum).'}`    
`r if(tech.memo){'2. Hook-and-line vessels with EM (*EM HAL* stratum).'}`
`r if(tech.memo){'3. Pot vessels (*OB POT* stratum).'}`
`r if(tech.memo){'4. Pot vessels with EM (*EM POT* stratum).'}`
`r if(tech.memo){'5. Trawl vessels (*OB TRW* stratum).'}`
`r if(tech.memo){'6. Trawl vessels participating in an exempted fishing permit (*EM TRW EFP* stratum).'}`

`r if(tech.memo){'The NMFS used only the trip-selection method (i.e., no vessel-selection) to assign observers and EM to vessels in the partial-coverage category for 2024.'}`

`r if(tech.memo){'In this report we attempt to evaluate the deployment of EM onto fixed gear vessels to the same degree as we evaluate the deployment of observers since catch accounting has used data collected through the *EM HAL* stratum since 2018 and the *EM POT* stratum since 2019. NMFS also sought vessels to participate in EM research and development activities. Vessels that volunteered for the fixed gear EM Program or EM research activities and were selected by the NMFS were not required to carry observers but were required to continue to log their fishing trips into the Observer Declare and Deploy System (ODDS).'}`

`r if(tech.memo){'The *EM TRW EFP* stratum is represented by vessels participating in an Exempted Fishing Permit (EFP) to evaluate the efficacy of EM on pollock catcher vessels using pelagic trawl gear in the Bering Sea and Gulf of Alaska. The EFP allows pollock catcher vessels using pelagic trawl gear to use EM systems *in lieu* of at-sea observers.  While we report some findings on these vessel activities here, we do not evaluate them fully sine the EFP is not part of the regulated fishery monitoring program.'}`

## The `r year` Annual Deployment Plan
The deployment design for the partial coverage component of the program involves three elements: (1) the selection method to accomplish random sampling; (2) division of the population of partial coverage trips/deliveries into selection strata; and (3) the allocation of deployment among strata.\
\
In `r year`, observers and EM were to be deployed using the trip-selection model in all ports throughout Alaska. Trip-selection refers to the method of selecting fishing trips as the sampling unit. Selection of trips was to be facilitated through vessel operators and owners logging their trips into the Observer Declare and Deploy System (ODDS) and being notified if the trip was selected for coverage.\
\
In 2024, NMFS implemented changes to both the stratification definitions and the allocation strategy (NMFS 2023). In previous years, strata were defined by the monitoring method (at-sea observer, EM, or none) and the gear type (either hook-and-line, pot, or trawl). The new strata definitions combined hook-and-line and/or pot gear trips together into a single fixed-gear type. Additionally, strata were further defined by the Fisheries Management Plan (FMP) the vessel intended to predominantly fish in — either the Bering Sea and Aleutian Islands (BSAI) or the Gulf of Alaska (GOA). The proximity allocation algorithm was utilized to determine the sampling rates of the partial coverage strata to reduce data gaps and impacts caused by small sample sizes. As an exception, trips by vessels participating in the walleye pollock (*Gadus chalcogrammus*, hereafter “pollock”) Trawl EM Exempted Fishing Permit (EFP) program were not included in the allocation algorithm due to  external funding sources for monitoring and its status as an experimental fishery.\
\
Vessels that join the fixed-gear EM program may log trips in the fixed-gear EM strata. Initiated by the Council in 2018, vessels that join the fixed-gear EM program use EM equipment to capture location and video information followed by human review for species identifications and counts. Weights for catch estimation are supplied from other sources.\
\
The Trawl EM EFP began in 2020, completed in 2024, and applied to trawl catcher vessels fishing with pelagic gear targeting pollock in both the BSAI and GOA. Vessels participating in this EFP used EM equipment to monitor for compliance of maximized retention of catch at sea and shoreside monitoring by observers during the delivery of catch. Vessels were able to opt-out of the EFP on a trip-by-trip basis and instead log trips in the at-sea observer strata. Trawl EM became a regulated monitoring program in 2025. Note that unlike other strata that use the trip as the PSU, the Trawl EM EFP strata instead defines the PSU as the shoreside delivery, as that is where the observer applies their sampling design.\
\
The deployment strata for `r year` (with abbreviation and coverage rate rounded to whole number) were defined as:

```{r, include = F}
strata.rates <- partial %>% 
  arrange(match(STRATA, strata.order)) %>%
  distinct(formatted_strat, Rate) %>%
  mutate(label = paste0(formatted_strat, " - ", round(Rate * 100, 0), "%"))
```
          
- Trips by fixed-gear vessels (using hook-and-line and/or pot gear) greater than or equal to 40 ft length overall (LOA) monitored with at-sea observers predominantly fishing in the Bering Sea and Aleutian Islands (`r strata.rates$label[1]`).\
- Trips by fixed-gear vessels greater than or equal to 40 ft LOA monitored with at-sea observers predominantly fishing in the Gulf of Alaska (`r strata.rates$label[2]`).\
- Trips by trawl vessels not participating in the Trawl EM EFP monitored with at-sea observers predominantly fishing in the Bering Sea and Aleutian Islands (`r strata.rates$label[3]`).\
- Trips by trawl vessels not participating in the Trawl EM EFP monitored with at-sea observers predominantly fishing in the Gulf of Alaska (`r strata.rates$label[4]`).\
- Trips by fixed-gear vessels listed under the Fixed-gear EM program monitored at-sea by EM systems and predominantly fishing in the Bering Sea and Aleutian Islands (`r strata.rates$label[5]`).\
- Trips by fixed-gear vessels listed under the Fixed-gear EM program monitored at-sea by EM systems and predominantly fishing in the Gulf of Alaska (`r strata.rates$label[6]`).\
- Shoreside deliveries by trawl vessels listed under the Trawl EM EFP (using pelagic trawl gear to target pollock) monitored 100% at-sea by EM systems, monitored shoreside by observers, and fishing in the Gulf of Alaska (`r sub("\\)", "]", sub("\\(", "[", strata.rates$label[7]))` shoreside). Note that vessels fishing under this EFP in the Bering Sea are managed within the full coverage category.\
- Fixed-gear vessels less than 40 ft LOA and vessels fishing with handline, jig, troll and dinglebar troll gear (*Zero coverage* - 0%).\

More information on the sampling design used by observers and the relationship between the sample design and catch estimation can be found in Cahalan and Faunce (2020) and the `r year` Observer Sampling Manual (AFSC `r year`). Bycatch estimates of Chinook salmon in the GOA are estimated using methods described in Cahalan et al. (2014). In the event that a delivery cannot be monitored (e.g., the case in a tendered delivery from a trip in an *OBS* strata or a non-pollock delivery), then estimation of bycatch comes by applying salmon bycatch rates to landed catch. Estimates of stock of origin from salmon bycatch are produced by the AFSC’s Auke Bay Laboratory. 
Although this chapter is focused on the partial coverage component of the fleet, the majority of the catch taken from the Federal waters off Alaska are completely monitored at the level of the trip (*Full coverage* - 100%). Vessels and processors in the full observer coverage category must comply with observer coverage requirements at all times when fish are harvested or processed. Specific requirements are defined in regulation at *50 CFR § 679.51(a)(2)*. The full coverage category includes the following:

- Catcher/processors (with limited exceptions);\
- Motherships;\
- Catcher vessels participating in programs that have transferable PSC allocations as part of a catch share program;\
- Catcher vessels using trawl gear that have requested placement in the full coverage category for all fishing activity in the BSAI for one year; and\
- Inshore processors receiving or processing Bering Sea pollock.\


## Performance Review Objectives  
`r if(tech.memo){'The following sections contain the FMSC review of the deployment of observers in'}` `r if(tech.memo){year}` `r if(tech.memo){'relative to the intended sampling plan and goals of the'}` `r if(tech.memo){year}` `r if(tech.memo){'ADP (NMFS'}` `r if(tech.memo){year-1}` `r if(tech.memo){'). This report identifies where potential mechanisms for biases exist and provides recommendations for further evaluation, including potential improvements to the observer deployment process that should be considered during the development of the'}` `r if(tech.memo){year + 2}` `r if(tech.memo){'ADP.'}`  
The following items from the `r year` ADP have been identified as objectives for evaluation in this report:

```{r actual_observerdays_paid, include=FALSE}

# Available budget for the year (in days)
OB_DAYS.BUDGET <- mean(bud_tbl$ADP_D)
  
#' Predicted cost for the year (in days)
OB_DOLLARS.BUDGET <- mean(bud_tbl$ADP_C)

# Actual number of observer days paid for
OB_DAYS.ACTUAL <- days_paid %>% 
  replace(is.na(.), 0) %>%
  summarise(Total.Days = sum(Base_Days + na.omit(Option_Days))) %>% 
  unlist() %>% 
  as.vector()

# Get total costs
OB_DOLLARS.ACTUAL <- days_paid %>% 
  summarise(Total.Base.Dollars = sum(Base_Dollars, na.rm = TRUE),
            Total.Option.Dollars = sum(Option_Dollars, na.rm = TRUE),
            Total.Travel.Dollars = sum(Travel_Dollars, na.rm = TRUE),
            Total.Quarantine.Dollars = sum(Quarantine_Dollars, na.rm = TRUE),
            Total.Plant.Dollars = sum(Plant_Dollars, na.rm = TRUE)) %>% 
  mutate(Total.Dollars = Total.Base.Dollars + Total.Option.Dollars + Total.Travel.Dollars + Total.Quarantine.Dollars + Total.Plant.Dollars)

# Percent of budget predicted to be spent
eval <- data.frame(OB_DAYS.ACTUAL, OB_DAYS.BUDGET) %>%
  mutate(
    # Percent of the day budget used
    pct.budget.spent = round(OB_DAYS.ACTUAL/OB_DAYS.BUDGET, digits = 3) * 100,
    # Percent of day budget remaining
    pct.budget.remain = round(100 - pct.budget.spent, digits = 3),
    # Was it more, less, or equal to the predicted budget?
    more.or.less = ifelse(pct.budget.remain > 0, " lower than ",
                          ifelse(pct.budget.remain == 0, "equal to",
                                 " greater than ")),
    quant.day = 100 * sum(OB_DAYS.ACTUAL >= bud_tbl$ADP_D) / length(bud_tbl$ADP_D),
    quant.dollars = 100 * sum(OB_DOLLARS.ACTUAL >= bud_tbl$ADP_C) / length(bud_tbl$ADP_C),
    dollars.perc = round(100 * (OB_DOLLARS.ACTUAL$Total.Dollars - OB_DOLLARS.BUDGET) / OB_DOLLARS.BUDGET, 1)
  )
```

1. [Deploy for the planned number of sea days specified in the `r year` ADP:]{.underline} This objective will be considered to be met if the actual number of sea days expended falls within the range of values from simulated sampling.\
2. [Deploy at the coverage rates specified in the `r year` ADP:]{.underline} For full and zero coverage, either the rate was equal to 100% or 0%, respectively. For strata under partial selection, coverage selection rates are expected to be within a 95% confidence interval computed from the realized coverage rates (under the assumption of a binomial distribution for observed trips).\
3. [Collect tissue samples from Chinook and chum (*Oncorhynchus keta*) salmon as specified in the `r year` Observer Sampling Manual to support the goal of collecting genetic samples from salmon caught as bycatch in groundfish fisheries to identify stock of origin:]{.underline} The sampling protocol established in the 2014 ADP (NMFS 2013, Faunce 2015) was used in `r year`. Under this protocol, observers on vessels delivering to shoreside processors in the GOA pollock trawl fishery monitor the delivery to enumerate salmon bycatch and obtain tissues for genetic analysis from the salmon bycatch. For trips in the *OB TRW GOA* stratum that are delivered to tender vessels and trips outside of the pollock fishery, observers obtain salmon counts and tissue samples from all salmon found within at-sea samples of the total catch. For the Trawl EM EFP, 100% of deliveries are monitored in the BSAI and 33.33% of deliveries in the GOA, including those from tenders.\
4. [Randomize deployment of observers into the partial coverage category of fishing activities:]{.underline} Evaluation of this objective is focused on the randomization of observer and EM deployments into primary sampling units, and how departures from a random sample affect data quality.      
  
## Observer Deployment Performance Metrics  
Performance metrics have been developed to assess whether the trip-selection process (through the implementation of the `r year` ADP) provides a representative sample of fishing trips in the North Pacific in `r year`. These metrics reflect four mechanisms that can impact the quality of the data: (1) sample frame discrepancies, (2) non-response, (3) differences in trip characteristics, and (4) sample size.\
\
The performance metrics used in this evaluation are as follows:

1. [Deployment rates for each stratum:]{.underline} This is the basic level of evaluation for comparing targeted and achieved sampling rates, where sampling strata are partitions of the entire population about which we want to make inferences (e.g., generate estimates of catch). Specifically, this section assesses the following:\
    a. Sample rates and number of samples relative to intended values.\
    b. Quantification of under- and over-coverage rates (sample frame discrepancies). Over-coverage of a population occurs when the sample frame includes elements that are not part of the target population. When these elements are included in the random sample, effort (i.e., time, cost) is expended needlessly. Under-coverage results from having a sample frame that does not include a portion of the target population which can lead to biased sampling if that portion of the population differs from the population included in the sample frame.\
    c. Non-response rates. Non-response occurs when randomly selected elements (trips or vessels) are not actually sampled. If these trips or vessels have different fishing behavior (e.g., catch, areas fished) than the rest of the population, the data collected will not represent the entire fleet (non-response bias).\
2. [Representativeness of the sample:]{.underline} Randomized sampling is a method used to ensure that the results of sampling reflect the underlying population. Departures from randomization can lead to non-representative data and hence potential bias in estimates of the parameters of interest. A randomized sample design is expected to achieve a rate of monitored events that is similar across both space and time. Representativeness of the sample was divided into four separate components:\
    a. Temporal representativeness. Plots of expected and actual monitoring rates over time, highlighting periods when these two rates deviate from each other are indicative of periods with differential realized sample rates (and potential temporal bias).\
    b. Spatial representativeness. Maps provide a visual depiction of the spatial distribution of monitoring coverage relative to effort in each partial coverage stratum, highlighting areas where more or fewer trips were monitored than expected.\
    c. Spatiotemporal distribution of coverage. The proportions of sample units monitored or nearby in time and space to monitored trips (the proximity indices) are compared to distributions of simulated outcomes to determine whether the realized coverage was distributed evenly in both time and space and whether the achieved coverage met the expectations of the selection rates prescribed by the ADP.\
    d. Representativeness of trip characteristics. Consistency of trip characteristics for monitored and unmonitored portions of the stratum. These metrics are based, in part, on the availability of data for both monitored and unmonitored fishing activities; for example, data that are reported for all trips on landing reports.\

Although these metrics can identify places where observed results differ from expectations, it is ultimately a subjective decision as to whether or not these differences are substantial enough to have management implications. This holds true even for tests that have associated *p*-values. 

# Changes to this report from last year
As mentioned previously, the strata definitions and allocation methods employed in `r year` differed from previous years. However, the evaluations performed in this chapter are largely unchanged.\

- Review timeliness for the *EM FIXED strata* is calculated in addition to data timeliness in order to account for the time between review and data availability that has been attributed to a NMFS database coding error.
- Monitoring rates for the *EM TRW (EFP)* strata are now performed at the level of the delivery, instead of at the level of the trip, as this more accurately represents how these strata are monitored. Trips that delivered their catch to tenders are also now accounted for and summarized as these deliveries are subject to monitoring by observers at shoreside processors. 
- Table 3-1 now includes the predicted and actual number of partial coverage trips to assess whether inaccuracies in the predicted fishing effort explains differences in predicted and actual monitored days.
- Summary tables for ODDS, Table 3-2 and Table 3-3, now separate counts of randomly selected trips and trips that inherited monitoring from previously selected and canceled trips. This helps to characterize the extent to which monitored trips are randomly selected or not.
- The analysis of spatial patterns were modified to use similar methods as the spatiotemporal analyses to better characterize whether different regions were over- or under-represented in the random samples, rather than the number of observed/reviewed trips in spatial cells.

# Evaluation of Deployment in `r year`  
The deployment of observers into the `r year` Federal fisheries off Alaska is primarily evaluated at the level of the deployment stratum because each stratum is defined by a different sampling rate or by a different monitoring method (e.g., observers versus EM). The ODDS is used to determine the strata of logged trips and randomly selects trips for monitoring at the rates prescribed by the `r year` ADP. Combined with predictions of fishing effort and monitoring costs, the ADP aims to maximize monitoring within budgetary constraints. However, the monitoring expenditures and number of monitored trips in each stratum that are actually realized may differ from the ADP's projections as a result of discrepancies between the expected and realized fishing effort in each stratum and/or the random nature of ODDS. This section compares the ADP's expectations with the realized outcomes.\


## Evaluating Effort Predictions  
```{r outcomes.plot, include=FALSE}

# Plot effort predictions against actual
outcomes.plot <- ggplot(bud_tbl, aes(x = ADP_D)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 20, alpha = 0.2, fill = "blue", col = "cornflowerblue") +
  geom_vline(xintercept = OB_DAYS.BUDGET, lty = 2, col = "black") +
  geom_vline(xintercept = OB_DAYS.ACTUAL, lty = 1, col = "blue") +
  annotate("text", x = OB_DAYS.BUDGET - 35, y = 0.002,
            label = paste("Available day budget: ", round(OB_DAYS.BUDGET), sep = ""),
            col = "black", angle = 90) + 
  annotate("text", x = OB_DAYS.ACTUAL - 35, y = 0.002,
           label = paste("Actual days paid: ", OB_DAYS.ACTUAL, sep = ""),
           col = "blue", angle = 90) +
  theme_bw() +
  theme(text = element_text(size = 14),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.line = element_line(linewidth = 0.5, color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_x_continuous(breaks = seq(2000, 3500, 500)) +
  #labs(y = paste("Density of simulated outcomes (", year - 1, "ADP)\n", sep = ""),
   #    x = paste("\nTotal days used in ", year - 1, sep = ""))
  labs(y = paste("Density of simulated outcomes"),
       x = paste("\nTotal days used"))

```

```{r costs.plot, include=FALSE}

# Plot predictions of Costs versus actual costs
costs.plot <- ggplot(bud_tbl, aes(x = ADP_C)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 20000, alpha = 0.2, 
                 fill = "green", col = "darkseagreen2") +
  geom_vline(xintercept = OB_DOLLARS.BUDGET, lty = 2, col = "black") +
  geom_vline(xintercept = OB_DOLLARS.ACTUAL$Total.Dollars, lty = 1, col = "darkgreen") +
  annotate("text", x = OB_DOLLARS.BUDGET - 45000, y = 0.0000015, 
           label = paste("Available contract budget: $", prettyNum(OB_DOLLARS.BUDGET, big.mark = ","), sep = ""), 
           col = "black", angle = 90) +
  annotate("text", x = OB_DOLLARS.ACTUAL$Total.Dollars - 45000, y = 0.0000015, 
           label = paste("Actual contract cost: $", prettyNum(OB_DOLLARS.ACTUAL$Total.Dollars, big.mark = ","), sep = ""), 
           col = "darkgreen", angle = 90) +
  theme_bw() +
  theme(text = element_text(size = 14),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.line = element_line(linewidth = 0.5, color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_x_continuous(labels = dollar) +
  #labs(y = paste("Density of simulated outcomes (", year - 1, " ADP", ")\n", sep = ""),
   #    x = paste("\nTotal contract costs in ", year - 1, sep = ""))
  labs(y = paste("Density of simulated outcomes"),
       x = paste("\nTotal contract costs"))

```

```{r budget_eval, include=FALSE}

#' *Table 3-1 : Comparison between predicted and actual at-sea trip days for partial coverage strata.*

# Evaluate why we were off budget
trip.budget.table <- 
  merge(
    # Predicted days by strata
    predicted %>%
      group_by(STRATA),
    # Actual trip days by strata
    merge(
      rbind(
        # EM strata
        work.data %>%
          filter(
            ADP == year & COVERAGE_TYPE == "PARTIAL" & STRATA %in% c("EM TRW GOA (EFP)", "EM FIXED BSAI", "EM FIXED GOA") & 
            OBSERVED_FLAG == "Y") %>%
          group_by(ADP, TRIP_ID, STRATA) %>%
          summarise(dmin = min(TRIP_TARGET_DATE, LANDING_DATE, na.rm = TRUE),
                    dmax = max(TRIP_TARGET_DATE, LANDING_DATE, na.rm = TRUE)) %>%
          mutate(days = ceiling(as.numeric(difftime(dmax, dmin, units = "days"))) + 1) %>%
          group_by(ADP, STRATA) %>% 
          summarise(act_days = sum(days)),
        # Observer strata
        filter(obs_act_days, ADP == year)), 
      work.data[ADP == year & COVERAGE_TYPE == "PARTIAL" & STRATA != "ZERO", .(act_N = uniqueN(TRIP_ID)), by = .(STRATA)],
      by = "STRATA"),
    all = TRUE) %>% 
  # Change NAs to 0 for actual days      
  mutate(act_days = ifelse(is.na(act_days), 0, act_days)) %>% 
  # Percent difference between predicted and actual
  mutate(diff = act_days -  pred_days,
         perc = (diff)/pred_days * 100,
         diff.N = act_N - pred_N,
         perc.N = diff.N / pred_N * 100) %>%
  mutate(perc = ifelse(is.infinite(perc), "-", format(perc, digits = 1, nsmall = 1)),
         perc.N = ifelse(is.infinite(perc.N), "-", format(perc.N, digits = 1, nsmall = 1))) %>%
  select(!c(ADP)) %>%
  arrange(factor(STRATA, levels = strata.order))

trip.budget.ob_total <- trip.budget.table %>%
  filter(STRATA %like% "OB") %>%
  group_by(STRATA = "At-sea observer total") %>%
  summarise(
    across(c(pred_N, pred_days, act_days, act_N, diff, diff.N), sum),
    perc.N = round(100 * sum(diff.N) / sum(pred_N), 1),
    perc = round(100 * sum(diff) / sum(pred_days), 1))
trip.budget.emfg_total <- trip.budget.table %>%
  filter(STRATA %like% "EM FIXED") %>%
  group_by(STRATA = "EM fixed-gear total") %>%
  summarise(
    across(c(pred_N, pred_days, act_days, act_N, diff, diff.N), sum),
    perc.N = round(100 * sum(diff.N) / sum(pred_N), 1),
    perc = round(100 * sum(diff) / sum(pred_days), 1))

# Append total
trip.budget.table <- 
  rbind(trip.budget.table %>% filter(STRATA %like% "OB"),
        trip.budget.ob_total,
        trip.budget.table  %>% filter(STRATA %like% "EM FIXED"),
        trip.budget.emfg_total,
        trip.budget.table  %>% filter(STRATA == "EM TRW GOA (EFP)"),
        trip.budget.table %>% 
          summarise(STRATA = "Total", 
                    pred_N = sum(pred_N),
                    act_N = sum(act_N),
                    diff.N = sum(diff.N),
                    perc.N = format(sum(diff.N)/sum(pred_N) * 100, digits = 1, nsmall = 1),
                    act_days = format(sum(act_days), digits = 1, nsmall = 0), 
                    pred_days = sum(pred_days), 
                    diff = sum(diff), 
                    perc = format(sum(diff)/sum(pred_days) * 100, digits = 1, nsmall = 1))) %>%
  mutate(pred_days = format(pred_days, digits = 1, nsmall = 0),
         diff = format(diff, digits = 1, nsmall = 0)) %>%
  select(STRATA, pred_N, act_N, diff.N, perc.N, pred_days, act_days, diff, perc)

# Cost per day comparison
# Expected CPD
cpd.budget <- OB_DOLLARS.BUDGET /  OB_DAYS.BUDGET
# Realized CPD
cpd.actual <- OB_DOLLARS.ACTUAL$Total.Dollars /  OB_DAYS.ACTUAL
# PERC DIFF - 
cpd.perc <- round((cpd.actual - cpd.budget) / cpd.budget * 100, 1)

```


Each year, NMFS sets an annual budget in the ADP for the deployment of partial coverage at-sea observers in terms of cost and observer days. The partial coverage observer budget for `r year` was set at \$`r formatC(bud_scen_lst, format = "d", big.mark = ",")` and `r formatC(OB_DAYS.BUDGET, format = "d", big.mark = ",")` days.\
\
In `r year`, FMA paid for `r formatC(round(OB_DAYS.ACTUAL), format = "d", big.mark = ",")` at-sea observer days, which was `r ifelse(eval$pct.budget.remain != 0, abs(eval$pct.budget.remain), " ")``r ifelse(eval$pct.budget.remain != 0, "%", "")` `r eval$more.or.less` the budgeted value and in `r round(eval$quant.day, 2)`th percentile of the simulated outcomes in the `r year` ADP (`r fig_nums('outcomes.costs.plot', display = "cite")`, top panel). However, the total number of fishing trips in the observed strata was only `r abs(trip.budget.ob_total$perc.N)`% lower than predicted, indicating that the lower-than-expected number of observed days was not due to fishing effort that was lower than anticipated (`r tbl_nums('trip.budget.tbl', display = "cite")`). Unsurprisingly, with fewer days observed than expected, the expenditures for partial coverage at-sea observers were also under budget by `r abs(eval$dollars.perc)`% (`r fig_nums('outcomes.costs.plot', display = "cite")`, bottom panel). The average cost of a partial coverage observer day in `r year` was only `r abs(cpd.perc)`% less than the cost that was estimated in the `r year` ADP.

In the *EM FIXED* strata, fishing effort by number of trips was `r trip.budget.emfg_total$perc.N`% higher than anticipated, but the number of sea days reviewed was `r abs(trip.budget.emfg_total$perc)`% lower than predicted. 

## Performance of the Observer Declare and Deploy System in Trip-Selection  
```{r odds.performance, warning=FALSE, include=FALSE}
#' Perform a check on the random number process in ODDS. `random.assign` is a flag for whether a trip's random number 
#' was less than the trip's selection rate. Since ODDS 3.0 went online in 2023, all trips have `RANDOM_NUMBER_USED` 
#' assigned and will therefore not have NA values. `PLANNED_EMBARK_DATE` is also changed from POSIX.ct to Date class.
#' Trips that inherited monitoring are not truly random, so such trips will 
odds.dat <- odds.dat %>% mutate(
  # Convert to Date class
  PLANNED_EMBARK_DATE = as.Date(PLANNED_EMBARK_DATE),
  # Compare random number to selection rate
  random.assign = case_when(
    is.na(RANDOM_NUMBER_USED) ~ NA,
    # Flag trips that inherited monitoring and therefore did not undergo random selection
    !is.na(INHERIT_TRIP_SEQ) ~ 100,
    # Trip was randomly selected to NOT be monitored OR not randomly selected (e.g., 100% compliance)
    (RANDOM_NUMBER_USED >= ODDS_SELECTION_PCT / 100) | (ODDS_SELECTION_PCT == 100) ~ 0,
    # Trip was randomly selected to be monitored
    RANDOM_NUMBER_USED < ODDS_SELECTION_PCT / 100 ~ 1,
    # Use -99 to identify records where none of the above conditions were met. 
    .default = -99
  )
)
if(nrow(odds.dat %>% filter(random.assign == -99))) stop("Some records with undefined 'random.assign'")

#' Add a flag for trips that were released (waived) from monitoring
odds.dat <- odds.dat %>%
  mutate(released = ifelse(!is.na(RELEASE_STATUS_DESCRIPTION) & RELEASE_STATUS_DESCRIPTION != "Release Denied", 1, 0))

# Are assignments being performed as programmed?
# TRIP_MONITOR_CODE
#   ND = Not observed (Provider Defaulted)
#   RO = Observer Trip (Require Observer)
#   OA = Observed Trip (Observer Assigned)
#   RL = Not Observed (Provider Release)
#   NO = Not Observed Trip (Observer Not Required or Waiver Granted)
# STRATA_CODE == 98 are EM trips that were marked "OA" only because they fished IFQ in more than one area
# STRATA_CODE == 96 are at-sea compliance trips that fished IFQ in more than one area
# Trip status codes: 
#   CS = Cancel by System 
#   PD = Pending
#   CN = Canceled 
#   CP = Completed 
#   CC = Cancel Cascaded (discontinued with ODDS 3.0 in 2023)
#   CR = Cancel Replaced (introduced with ODDS 3.0 in 2023)

#' Based on a trip's stratum and the outcome of the trip, flag trips trips that were either monitored or intended to be
#' monitored as a result of random selection in `random.test`. Trips whose fate was predetermined by prior trips, e.g., 
#' via inherits and cascade cancels OR were undetermined, e.g., system canceled, OR requested monitoring when fishing 
#' IFQ/CDQ in multiple areas are excluded with a value of [NA]. 
#' Use `TRIP_MONITOR_CODE` to identify trips in non-100% compliance monitoring strata that were slated for monitoring, 
#' whether it was realized or not, including those where the trip was waived [1]. Finally, trips in strata without 
#' random selection and trips marked as not monitored are identified [0].
odds.dat <- odds.dat %>% mutate(
  random.test = case_when(
    #' `Trips that did not go through random selection [NA]`
    # Exclude trips that inherited other trips (selection of this record was not random)
    !is.na(INHERIT_TRIP_SEQ) ~ NA, 
    # Exclude trips missing random numbers or cancel cascaded
    TRIP_STATUS_CODE == "CC" | is.na(RANDOM_NUMBER_USED) ~ NA,
    # Exclude trips canceled by system, which have an unknown monitoring outcome and do not become inherits
    TRIP_STATUS_CODE == "CS" ~ NA,
    # Exclude trips that requested coverage (multi-area IFQ/CDQ trips)
    USER_REQST_COVERAGE == "Y" ~ NA,
    
    #' `Trips that did go through random selection and were selected [1]`
    # Trip was randomly selected for monitoring but was released
    released == 1 ~ 1,
    # Trip was monitored or supposed to be monitored due to random selection (not within compliance monitoring)
    TRIP_MONITOR_CODE %in% c("ND", "RO", "OA", "RL") & ODDS_SELECTION_PCT != 100 ~ 1,
    
    #' `Trips that did go through random selection and were NOT selected [0]`
    # Trip was not monitored because it was not selected OR the trip was not randomly selected because it had 100% compliance monitoring
    (TRIP_MONITOR_CODE == "NO") | ODDS_SELECTION_PCT == 100 ~ 0,  
    .default = -99
  )
)
if(nrow(odds.dat %>% filter(random.test == -99))) stop("Some records with undefined 'random.test'")

#' Check whether the monitoring outcome matches what was determined by random selection, i.e., trips that were not monitored when they were supposed to be or vice versa. Trips that did not under random selection will return NA.
odds.dat <- odds.dat %>% mutate(random.test.result = ifelse(random.test - random.assign == 0, "PASS", "FAIL"))

#' \TODO The check below had two vessels that failed the check, and in both instances they were OB FIXED BSAI stratum trips that had three-in-a-row releases. We had a known bug that the release table was not populating when ODDS correctly applied the non-selected status to these trips. GMM added this issue to P-Track. For Now I'm hardcoding this fix.
if(year == 2024) {
  cat(paste("2024 only: hard-coding releases for two trips that failed the `random.test.result` due to three-in-a-row releases in odds.odds_strata_release not populating correctly."))
  setDT(odds.dat)
  odds.dat[TRIP_PLAN_LOG_SEQ %in% c(202432902, 202427490), 
           ':=' (TRIP_MONITOR_CODE = "RL", RELEASE_COMMENT = "Three-in-a-row", 
                 RELEASE_STATUS_DESCRIPTION = "Three-in-a-row", random.test.result = "PASS", released = 1)]
  setDF(odds.dat)
}
if(nrow(filter(odds.dat, random.test.result == "FAIL"))) {
  stop(paste0("There are trips that failed `random.test.result`!"))
} else {
  odds.dat <- odds.dat %>% select(-c(random.test, random.test.result))
}
```

```{r disposition.tbl, warning=FALSE, include=FALSE}

#' Disposition of trips in the ODDS. This table summarizes the frequency of trip activities.
#' It does not represent the actual trips expected on the grounds. That is the next table.
#' This was Table 3-2 in the 2019 Annual Report, but has not been generated since then. The data.frame is still 
#' referenced in the text.

# Summarize by stratum and random.assign
disposition <- odds.dat %>%
  filter(ODDS_SELECTION_PCT != 100 & (USER_REQST_COVERAGE %in% c("N", NA)) ) %>%
  distinct(YEAR, STRATA, random.assign, TRIP_STATUS_CODE, TRIP_PLAN_LOG_SEQ, released, TRIP_PLAN_NUMBER) %>%
  group_by(YEAR, STRATA, random.assign) %>%
  summarize(Trips.logged.ODDS = length(unique(TRIP_PLAN_LOG_SEQ)),
            Trips.canceled.bysystem = length(TRIP_PLAN_LOG_SEQ[TRIP_STATUS_CODE == "CS"]),
            Trips.remaining = Trips.logged.ODDS - Trips.canceled.bysystem,
            Trips.canceled.byusers = length(TRIP_PLAN_LOG_SEQ[TRIP_STATUS_CODE %in% c("CN", "CC", "CR")]),
            Paper.tickets = length(TRIP_PLAN_LOG_SEQ[!is.na(TRIP_PLAN_NUMBER)])) 

# Add yearly totals and summary statistics
disposition <- merge(
  disposition,
  disposition %>% 
    group_by(YEAR, random.assign) %>% 
    summarise(STRATA = "Total", across(where(is.integer), sum)) %>%
    data.frame(),
  all = TRUE) %>%
  mutate(Cancellation.Pct = format(round((Trips.canceled.byusers / Trips.remaining) * 100, 1), nsmall = 1)) %>%
  # Redefine values in random.assign
  mutate(random.assign = factor(case_when(is.na(random.assign) | random.assign == 1 ~ "Random",
                                          random.assign == 0 ~ "Not selected",
                                          random.assign == 100 ~ "Inherited"),
                                levels = c("Not selected", "Random", "Inherited")))

# Then rename some columns for the table
# Table 3-2 (2019, cancellation rates AFTER random selection) ---- 

disposition.table <- disposition %>%
  arrange(YEAR, match(STRATA, strata.order)) %>%
  rename("Strata" = STRATA,
         "Year" = YEAR,
         "Selection outcomes" = random.assign,
         "Logged (*a*)" = Trips.logged.ODDS,
         "Canceled by system (*b*)" = Trips.canceled.bysystem,
         "Trips remaining (*c* = *a*-*b*)" = Trips.remaining,
         "Canceled by user (*d*)" = Trips.canceled.byusers,
         "Paper" = Paper.tickets,
         "% User cancellation (*d*/*c* x 100)" = Cancellation.Pct)

#' So far, the vessels that requested monitoring, i.e., multi-area IFQ/CDQ trips, have been excluded in these counts because these trips are not subject to random selection. Count the number of completed trips that requested monitoring by stratum.
#' \TODO incorporate the `multi_area_ifq.string` object in the text to list off which trips did not undergo random selection.
multi_area_ifq <- odds.dat %>% 
  filter(USER_REQST_COVERAGE == "Y" & TRIP_STATUS_CODE %in% c("CP", "PD")) %>%
  group_by(YEAR, STRATA) %>%
  summarize(REQ = uniqueN(TRIP_PLAN_LOG_SEQ)) %>%
  arrange(match(STRATA, strata.order))
multi_area_ifq_vec <- apply(multi_area_ifq, 1, function(x) paste0(
    x[['REQ']], " trip", if(x[['REQ']] > 1) "s", " in the ", x[['STRATA']], " stratum"))
multi_area_ifq_vec.length <- length(multi_area_ifq_vec)
if(multi_area_ifq_vec.length > 2) {
  multi_area_ifq.string <- paste0(
    paste0(multi_area_ifq_vec[-multi_area_ifq_vec.length], collapse = ", "), " and ",
    multi_area_ifq_vec[multi_area_ifq_vec.length])
} else if(mutli_arae_ifq_vec == 2) {
  multi_area_ifq.string <- paste0(multi_area_ifq_vec[1], " and ", multi_area_ifq_vec[2])
} else multi_area_ifq.string <- multi_area_ifq_vec
rm(multi_area_ifq, multi_area_ifq_vec, multi_area_ifq_vec.length)

# Logged trip summaries
odds.logged <- disposition.table %>% filter(Strata == "Total") %>% 
  summarize(total = sum(`Logged (*a*)`),
            sel.n = `Logged (*a*)`[`Selection outcomes` == "Not selected"],
            sel.r = `Logged (*a*)`[`Selection outcomes` == "Random"],
            sel.i = `Logged (*a*)`[`Selection outcomes` == "Inherited"],
            can.sys = sum(`Canceled by system (*b*)`),
            can.sys.p = round(100 * can.sys / total, 1),
            can.usr = sum(`Canceled by system (*b*)`),
            can.usr.p = round(100 * can.usr / total, 1),
            can.total = can.sys + can.usr,
            can.total.p = round(100 * can.total / total, 1),
            ) %>% 
  prettyNum(big.mark = ",")
odds.can.usr.prop <- disposition.table %>% filter(Strata != "Total") %>%
  group_by(Strata) %>%
  reframe(r_n_prop = as.numeric(`% User cancellation (*d*/*c* x 100)`)[`Selection outcomes` == "Random"] /
            as.numeric(`% User cancellation (*d*/*c* x 100)`)[`Selection outcomes` == "Not selected"] ) %>%
  filter(!is.infinite(r_n_prop)) %>%
  slice_max(r_n_prop)

```

```{r odds.time_series, warning=FALSE, include=FALSE}

#' Create a time series of trips in ODDS by stratum where the counts of randomly-selected, inherited, and waived trips
#' are totaled for each day. These time series will be totaled for `odds.logged.tbl` and converted to cumulative totals
#' in`odds.ts.daily.cuml`

#' Fist, subset the data to exclude trips where random selection is irrelevant, i.e., zero selection, 100% selection, and trips with voluntary monitoring. Ensure 1 row = 1 TRIP_ID
odds.ts.sub <-  odds.dat %>%
  filter(!ODDS_SELECTION_PCT %in% c(0, 100) & (USER_REQST_COVERAGE %in% c("N", NA))) %>%
  mutate(Date = as.Date(PLANNED_EMBARK_DATE, format = "%Y-%m-%d")) %>% 
  # Simplify the dataset such that each record is one trip
  select(YEAR, Date, TRIP_PLAN_LOG_SEQ, TRIP_STATUS_CODE, ODDS_SELECTION_PCT, RANDOM_NUMBER_USED, 
         INHERIT_TRIP_SEQ, STRATA, random.assign, released) %>%
  unique() %>% 
  # Group the data by stratum and date for subsequent summaries
  group_by(YEAR, STRATA, Date) 
if(any(duplicated(odds.ts.sub$TRIP_PLAN_LOG_SEQ))) stop("There are duplicated TRIP_ID in odds.ts.sub!")

#' Count trips subject to random selection BEFORE cancellations, inherits and waivers
odds.ts.all <- odds.ts.sub  %>% 
  filter(is.na(INHERIT_TRIP_SEQ)) %>%
  summarize(
    # Count number of logged trips that were randomly selected on each date
    n = sum(random.assign, na.rm = TRUE),
    N = n(),
    TYPE = "Original date: All logged trips") %>% 
  arrange(Date) %>% 
  ungroup()

#' Count completed or pending trips only, including counts for inherits and waivers
odds.ts.cp <- odds.ts.sub %>% 
  filter(TRIP_STATUS_CODE %in% c("CP", "PD")) %>%
  summarize(
    # Count trips subject to random selection
    n = sum(random.assign[is.na(INHERIT_TRIP_SEQ)], na.rm = TRUE),
    # Count trips that inherited selection
    ni = sum(!is.na(INHERIT_TRIP_SEQ)),
    # Count trips that were released
    n.r = sum(released[is.na(INHERIT_TRIP_SEQ)], na.rm = TRUE),
    ni.r = sum(released[!is.na(INHERIT_TRIP_SEQ)], na.rm = TRUE),
    N = n(),
    TYPE = "Adjusted date: Valid trips only") %>% 
  arrange(Date) %>% 
  ungroup()

#' Get the dates that inherited trips were created and realized, i.e., canceled randomly-selected trips where coverage 
#' was not released
odds.ts.ni.create <- odds.ts.sub %>% 
  filter(!(TRIP_STATUS_CODE %in% c("CP", "PD")) & is.na(INHERIT_TRIP_SEQ) & released == 0) %>%
  summarize(ni.create = sum(random.assign, na.rm = TRUE)) %>%
  filter(ni.create > 0) %>%
  mutate(TYPE = "Original date: All logged trips") %>%
  ungroup()

#' Count inherited trips that were completed and not released
odds.ts.ni.real <- odds.ts.sub %>% 
  filter(TRIP_STATUS_CODE %in% c("CP", "PD") & !is.na(INHERIT_TRIP_SEQ) & released == 0) %>%
  summarize(ni.real = n()) %>% 
  filter(ni.real > 0) %>%
  mutate(TYPE = "Adjusted date: Valid trips only") %>%
  ungroup()

rm(odds.ts.sub)
```

```{r odds.logged.tbl, warning=FALSE, include=FALSE}

#' \TODO 'logged' is perhaps not the best label as all trips are logged, and should instead specify that these are completed/closed trips. The table caption for Table 3-3 could also better describe this.
#' Summary of ODDS *completed trips* disposition by stratum, with total completed trips, total randomly selected 
#' trips, total inherited trips, waived trips from either group, and the resulting totals.

logged <- odds.ts.cp %>%
  group_by(STRATA) %>%
  summarize(Total.trips = sum(N), 
            Random.Selected = sum(n), 
            Inherit.Selected = sum(ni),
            Random.Waived = sum(n.r), 
            Inherit.Waived = sum(ni.r)) %>%
  melt(id.vars = c("STRATA", "Total.trips")) %>%
  mutate(GROUP = factor(case_when(variable %like% "Random." ~ "Random",
                                  variable %like% "Inherit." ~ "Inherit"),
                        levels = c("Random", "Inherit")),
         variable = sub("Random.|Inherit.", "", variable)) %>%
  dcast(STRATA + GROUP + Total.trips ~ variable, value.var = "value", fill = 0) %>%
  group_by(STRATA) %>%
  mutate(Total.final.selected = sum(Selected - Waived),
         Pct.selected.from.inherit  = round(100 * (Selected[GROUP == "Inherit"] - Waived[GROUP == "Inherit"]) / Total.final.selected, 1),
         Pct.reduced.from.waiver   = round(100 * sum(Waived) / (Total.final.selected + sum(Waived)), 1)) %>%
  ungroup() %>%
  arrange(match(STRATA, strata.order))
  
# Inherits and release rates AFTER random selection and cancellation) ---- 

# Tells us with more granularity how many selected trips were canceled or waived. 

# Disposition of trips - showing only valid trips and categories are exclusive
logged.table <- logged %>%
  bind_rows(
    logged %>% 
      summarise(STRATA = "Total",
                Total.trips = sum(Total.trips)/2,
                Total.inherited = sum(Selected[GROUP == "Inherit"]),
                Selected = sum(Selected),
                Waived = sum(Waived),
                Total.final.selected = Selected - Waived,
                Pct.selected.from.inherit = round(100 * Total.inherited / Total.final.selected, 1),  
                Pct.reduced.from.waiver   = round(100 * Waived / (Total.final.selected - Waived), 1)) %>% 
      select(-Total.inherited)) %>%
  select("Strata" = STRATA, 
         "Total completed trips" = Total.trips,
         "Selection type" = GROUP,
         "Selected" = Selected, 
         "Waived" = Waived, 
         "Total final selected" = Total.final.selected,
         "% Selected from inherits" = Pct.selected.from.inherit, 
         "% Reduction of selected trips due to waivers" = Pct.reduced.from.waiver) %>% 
  arrange(match(Strata, strata.order))
```

```{r selection.tbl, warning=FALSE, include=FALSE}
# Table 3-4 (2019) or Table 3-2 (2020) (Full ODDS summary, realized versus expected rates) ----
# This table will show probabilities that selection rates = programmed rates under various assumptions

for.p.wide <- merge(
  # Completed non-100% monitored trips only (after cancellations)
  logged %>% 
    group_by(STRATA) %>%
    summarize(random.n = Selected[GROUP == "Random"],
              # after inherits
              inherit.n = sum(Selected - Waived),
              #'total before waivers*
              inherit.butnowaiver.n = sum(Selected),
              Total.n = unique(Total.trips),
              Source = "Valid trips only"),
  #' Now add in the total trips for the random number only. This includes every trip logged in ODDS (excluding those in 
  #' 100% selection) and counts how many were selected based on the random number and selection rate, `before`
  #' inherits, cancellations, and waivers
  odds.ts.all %>%
    select(-YEAR) %>%
    group_by(STRATA) %>%
    summarise(random.n = sum(n), Total.n = sum(N), Source = "ODDS random number"),
  all = TRUE) %>%
  arrange(Source, STRATA)

# Merge in the selection rates of each stratum for calculating p-values later
for.p.wide <- for.p.wide %>%
  left_join(
    partial %>% select(-GEAR, -YEAR) %>% distinct(), 
    by = c("STRATA")) %>%
  mutate(ODDS_SELECTION_PCT = Rate * 100) %>%
  select(!c(formatted_strat, Rate))

# Put into long format
p.table <- for.p.wide %>%
  pivot_longer(cols = c(-STRATA, -Total.n, -ODDS_SELECTION_PCT, -Source), names_to = "key", values_to = "value")

# Omit records with missing values
p.table <- p.table[!is.na(p.table$value), ]
p.table$p <- NA_real_
p.table$pct <- NA_real_

# Run the binomial test for each  row to see if realized rates met expectation
for(i in 1:nrow(p.table)){
  # Realized monitoring rate
  p.table$pct[i] = round((p.table$value[i] / p.table$Total.n[i]) * 100, 2)
  # Check out this format - forces digits even if zero value!
  p.table$p[i] = format(round(stats::binom.test(p.table$value[i], p.table$Total.n[i], 
                        p = p.table$ODDS_SELECTION_PCT[i] / 100, 
                        alternative = "two.sided")$p.value, 3), nsmall = 3)
}

#' Format programmed and realized monitoring rates as percentages to 2 decimal places.
p.table <- p.table %>% 
  mutate(ODDS_SELECTION_PCT = round(ODDS_SELECTION_PCT, 2), 
         pct = round(pct, 2))

# Cleanup
rm(for.p.wide)

# rename values for one column
p.table$key[p.table$key == "inherit.n"] <- "Expected"
p.table$key[p.table$key == "random.n"] <- "Random Selection Only"
p.table$key[p.table$key == "inherit.butnowaiver.n"] <- "Expected if no waivers"

# Create table
Selection.table <- p.table %>%
  mutate(p = ifelse(p < 0.05, paste0(p, "*"), p)) %>%
  arrange(STRATA, Source, desc(key)) %>%
  arrange(match(STRATA, c("OB HAL", "EM HAL", "OB POT", "EM POT", "OB TRW"))) %>% 
  mutate(new_trip_type = 
        derivedFactor("Initial Random Selection, *a*"= Source == "ODDS random number",
                      "After Cancellations, *b* (*a*-*b*)" = key == "Random Selection Only" & Source == "Valid trips only",
                      "With Inherits, *c* (*a*-*b*+*c*)" = key == "Expected if no waivers",
                      "After Waivers, *d* (*a*-*b*+*c*-*d*)" = key == "Expected",
                      .ordered = TRUE, .method = "unique")) %>% 
  select("Strata" = STRATA, 
         "Trip disposition" = new_trip_type,   # UPDATE THESE NAMES
         "n" = value,  # "Selected trips"
         "N" = Total.n,   # "Total trips"
         "Actu" = pct,   # "Actual selection (%)"
         "Prog" = ODDS_SELECTION_PCT,  # "Programmed selection (%)"
         "*p*" = p) %>%    # "*p*-value (H0: Actual = Programmed)"
  arrange(factor(Strata, levels = strata.order))
```

```{r odds.ts.daily.cuml.fig, warning=FALSE, include=FALSE}

# Use the ODDS time series to plot if and when the realized monitoring rates are within the expected range given the
# programmed rates and the binomial distribution. Calculate running totals of trips and monitored trips.
odds.ts.daily.cuml <- bind_rows(
  # Initial random selection
  odds.ts.all, 
  # Final after cancellations, inherits, and waivers
  odds.ts.cp %>% 
    group_by(YEAR, STRATA, Date, TYPE) %>%
    summarise(n = n + ni - n.r - ni.r, N = N)
  ) %>% 
  group_by(YEAR, STRATA, TYPE) %>%
  mutate(
    TYPE = as.factor(TYPE),
    running.n = order_by(Date, cumsum(n)),
    running.N = order_by(Date, cumsum(N)),
    running.pct = round((running.n / running.N) * 100, 2)
  ) %>%
  arrange(STRATA, TYPE, Date) %>%
  right_join(
    partial %>%
      filter(STRATA != "EM TRW GOA (EFP)") %>%
      distinct(YEAR, STRATA, ODDS_SELECTION_PCT = Rate * 100),
    by = c("STRATA", "YEAR")) %>%
  ungroup()

# reorder strata for plot
odds.ts.daily.cuml <- transform(odds.ts.daily.cuml,
                                STRATA = factor(STRATA, levels = c(
                                  "OB FIXED BSAI", "OB FIXED GOA", "OB TRW BSAI", "OB TRW GOA", 
                                  "EM FIXED BSAI", "EM FIXED GOA")),
                                YEAR = as.character(year))

# Generate data for ribbon plots for the random data only.
odds.ts.daily.cuml$L95 <- ifelse(
                          odds.ts.daily.cuml$TYPE == "Adjusted date: Valid trips only", 
                          NA, 
                          round((qbinom(0.025, 
                                        odds.ts.daily.cuml$running.N, 
                                        odds.ts.daily.cuml$ODDS_SELECTION_PCT / 100)) / odds.ts.daily.cuml$running.N, 3))

odds.ts.daily.cuml$H95 <- ifelse(
                          odds.ts.daily.cuml$TYPE == "Adjusted date: Valid trips only", 
                          NA, 
                          round((qbinom(0.975,
                                        odds.ts.daily.cuml$running.N,
                                        odds.ts.daily.cuml$ODDS_SELECTION_PCT / 100)) /
                                        odds.ts.daily.cuml$running.N, 3))

# Calculate the p-values that the realized rates are different from the expected (programmed) rate
for(i in 1:nrow(odds.ts.daily.cuml)){
  odds.ts.daily.cuml$test_stat_random[i] <- ifelse(
    odds.ts.daily.cuml$running.n[i] < 1, NA,
    round(binom.test(odds.ts.daily.cuml$running.n[i],
                     odds.ts.daily.cuml$running.N[i],
                     odds.ts.daily.cuml$ODDS_SELECTION_PCT[i] / 100,
                     alternative = "two.sided")$p.value, 3))
}

rm(i)

for.plot.text <- left_join(
  odds.ts.daily.cuml %>% 
    group_by(YEAR, STRATA, TYPE) %>%
    filter(Date == max(Date)) %>%
    group_by(YEAR, STRATA) %>%
    mutate(label.original = format(running.pct[TYPE == "Original date: All logged trips"], digits = 4, nsmall = 2),
          label.adjusted = format(running.pct[TYPE == "Adjusted date: Valid trips only"], digits = 4, nsmall = 2)) %>% 
    distinct(YEAR, STRATA, label.original, label.adjusted),
  partial %>% 
    filter(STRATA != "EM TRW EFP") %>%
    select(YEAR, STRATA, label.programmed = Rate) %>%
    mutate(STRATA = as.character(STRATA), 
          YEAR = as.factor(YEAR),
          label.programmed = format(label.programmed * 100, digits = 4, nsmall = 2)), 
  by = c("STRATA", "YEAR"))

odds.ts.daily.cuml <- merge(odds.ts.daily.cuml, for.plot.text)

# Custom y axes for each row of the figure
odds.scales <- list(
  scale_y_continuous(limits = c(20, 70), breaks = c(0, 20, 40)),
  scale_y_continuous(limits = c(0, 30), breaks = c(0, 20, 40)),
  scale_y_continuous(limits = c(40, 100)),
  scale_y_continuous(limits = c(0, 40)),
  scale_y_continuous(limits = c(50, 100)),
  scale_y_continuous(limits = c(0, 75))
)

odds.by.date.plot <- 
  ggplot(filter(odds.ts.daily.cuml, YEAR == year), aes(x = Date, col = fct_rev(TYPE))) +
  facet_grid(factor(STRATA, levels = strata.order) ~., scales = "free") +
  # Add lines for selection rates  
  geom_hline(aes(yintercept = ODDS_SELECTION_PCT), lty = 3, linewidth = 1) +
  geom_ribbon(aes(ymin = L95 * 100, ymax = H95 * 100), alpha = 0.1) +
  geom_line(aes(y = running.pct), linewidth = 1.5) +
  scale_color_manual(name = "Rate Type", values = c("gray50", "black")) +
  ylab("Cumulative coverage percent\n") +
  xlab("\nDate") +
  # Bottom rug lines indicate when selected trips were canceled, i.e. when inherits are created 
  geom_rug(data = odds.ts.ni.create %>% filter(YEAR == year), sides = "b", color = "gray50") + 
  # Top rug lines indicate when inherited monitoring is realized on a completed trip
  geom_rug(data = odds.ts.ni.real %>% filter(YEAR == year), sides = "t") + 
  geom_text(aes(x = as.Date(Inf), y = Inf, label = paste("Programmed (%) = ", label.programmed, sep = "")), 
            hjust = 1.02, vjust = 2, size = 3, color = "black", check_overlap = TRUE) +
  geom_text(aes(x = as.Date(Inf), y = Inf, label = paste("Original (%) = ", label.original, sep = "")), 
            hjust = 1.02, vjust = 3.2, size = 3, color = "gray50", check_overlap = TRUE) +
  geom_text(aes(x = as.Date(Inf), y = Inf, label = paste("Final (%) = ", label.adjusted, sep = "")), 
            hjust = 1.02, vjust = 4.4, size = 3, color = "black", check_overlap = TRUE) +
  fig.theme + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90),
        strip.text.x = element_text(size = 10),
        legend.position = "none") +
  ggh4x::facetted_pos_scales(y = odds.scales)

```

```{r odds_paragraph, warning=FALSE, include=FALSE}

#' \TODO Not currently used but could/should be?

# Objects that help increase readability of in-text calculations in the following paragraphs

# Total canceled trips (both by ODDS or by users)
totl.canc <- disposition %>% 
  filter(STRATA == "Total" & random.assign == "Selected", YEAR == year) %>%
  summarize(totl.canc = sum(as.numeric(Trips.canceled.bysystem)) + sum(as.numeric(Trips.canceled.byusers))) %>% 
  as.numeric()

# Percent of total trips canceled by the ODDS system
perc.syst.canc <- disposition %>% 
   filter(STRATA == "Total" & random.assign == "Selected", YEAR == year) %>% 
   summarize(
     perc.syst.canc = round(
       (sum(as.numeric(Trips.canceled.bysystem)) / sum(as.numeric(Trips.logged.ODDS))) * 100, 1)) %>% 
   as.numeric()

#Percent of total trips canceled by users
perc.user.canc <- disposition %>% 
  filter(STRATA == "Total" & random.assign == "Selected", YEAR == year) %>% 
  summarize(perc.syst.canc = format(
    (sum(as.numeric(Trips.canceled.byusers)) / sum(as.numeric(Trips.logged.ODDS))) * 100, digits = 1, nsmall = 1)) %>% 
  as.character()

# Lowest and highest cancellation rate/stratum for selected trips by strata
sel.min.canc <- disposition %>% 
                filter(STRATA != "Total" & random.assign == "Selected") %>%
                group_by(YEAR, STRATA) %>% 
                summarize(Cancellation.Pct = (sum(as.numeric(Trips.canceled.byusers)) / sum(as.numeric(Trips.remaining))) * 100) %>%
                slice_min(as.numeric(Cancellation.Pct)) %>% 
                mutate(Cancellation.Pct = round(Cancellation.Pct, 1))

sel.max.canc <- disposition %>% 
                filter(STRATA != "Total" & random.assign == "Selected") %>%
                group_by(YEAR, STRATA) %>% 
                summarize(Cancellation.Pct = (sum(as.numeric(Trips.canceled.byusers)) / sum(as.numeric(Trips.remaining))) * 100) %>%
                slice_max(as.numeric(Cancellation.Pct)) %>% 
                mutate(Cancellation.Pct = round(Cancellation.Pct, 1))
```

```{r odds_paragraph2, warning=FALSE, include=FALSE}

#' \TODO Not currently used but could/should be?
 
# Lowest and highest percentages of valid trips that were selected from the inherit process.
inherit_waiver.smry <- logged %>% distinct(STRATA, Pct.selected.from.inherit, Pct.reduced.from.waiver)

inherit.min <- inherit_waiver.smry %>% 
  slice_min(as.numeric(Pct.selected.from.inherit))  %>% 
  select(-Pct.reduced.from.waiver) 

inherit.max <- inherit_waiver.smry %>% 
  slice_max(as.numeric(Pct.selected.from.inherit))  %>% 
  select(-Pct.reduced.from.waiver) 

waiver.min <- inherit_waiver.smry %>% 
  slice_min(as.numeric(Pct.reduced.from.waiver))  %>% 
  select(-Pct.selected.from.inherit) 

waiver.max <- inherit_waiver.smry %>% 
  slice_max(as.numeric(Pct.reduced.from.waiver))  %>% 
  select(-Pct.selected.from.inherit) 
```

```{r odds_paragraph3, warning=FALSE, include=FALSE}

#' \TODO Not currently used but could/should be?

# Text strings for reporting on the observer initial selection rates by strata
pct_txt <- p.table %>% 
           filter(Source == "ODDS random number") %>% 
           mutate(pct_txt = paste0(pct, "% for the *", STRATA, "* stratum")) %>%
           distinct(pct_txt) 

# Extract p-values associated with random selection (H0: observed = expected
# rates). Merge to the partial look up table to ensure that things stay in order.
random_pvals <- merge(
  p.table, 
  partial %>% 
    filter(STRATA != "EM TRW EFP") %>% 
    distinct(STRATA, formatted_strat)) %>% 
  filter(Source == "ODDS random number") %>% 
  select(p, formatted_strat)
 
# Text strings for reporting on the observer FINAL selection rates by strata
final_pct_txt <- p.table %>% 
                 filter(key == "Expected") %>% 
                 mutate(pct_txt = paste0(pct, "% for the *", STRATA, "* stratum")) %>% 
                 distinct(pct_txt) 

# Extract p-values associated with FINAL selection (H0: final observed = expected
# rates).
final_pvals <- merge(
  p.table, 
  partial %>% 
    filter(STRATA != "EM TRW EFP") %>% 
    distinct(STRATA, formatted_strat)) %>% 
  filter(key == "Expected") %>% 
  select(p, formatted_strat)
```


The ODDS facilitates the random selection for monitoring in strata and fishers are required to log anticipated fishing trips. The ODDS generates a random number according to the programmed rates from the ADP and assigns each logged trip to either “selected to be monitored” (selected) or “not selected to be monitored” (not selected) categories. The ODDS is not used to select which deliveries are to be monitored by shoreside observers for the *EM TRW GOA (EFP)* stratum which is excluded from the following summaries. In addition, trips that requested monitoring to fulfill monitoring requirements when fishing CDQ/IFQ in multiple reporting areas were excluded from this analysis as such trips are not subject to random sampling by ODDS.\
\
Logged trips have different dispositions. When initially logged, trips are considered pending, and subsequently have two dispositions: closed or canceled. A trip can be closed by (1) selecting landing reports from a menu or (2) manually entering the end of trip information for observed trips. The vessel operator may change the dates of a logged trip regardless of selection status prior to, or in lieu of, cancellation. However, trips that have not been closed at the end of the calendar year are automatically canceled by ODDS ('canceled by system') to prevent `r year` ODDS trips from affecting the deployment rates set for the `r year + 1` ADP year. Trips that were selected to be monitored by ODDS that are subsequently canceled trigger the next newly-logged trip (for at-sea observers) or next logged trip (for fixed-gear EM) to automatically inherit the selected status. These trips are termed inherited trips. Note that beginning in 2025, vessel operators in the observed strata are no longer able to cancel logged trips.\
\
The number of trips logged in ODDS in `r year` and their dispositions is summarized in `r tbl_nums('disposition.table', display = "cite")`. Of the `r odds.logged[["total"]]` total trips logged, `r odds.logged[["sel.r"]]` were randomly selected and `r odds.logged[["sel.i"]]` inherited monitoring from a previously canceled selected trip. In total, `r odds.logged[["can.total"]]` (`r odds.logged[["can.total.p"]]`%) were canceled: `r odds.logged[["can.sys"]]` by ODDS (`r odds.logged[["can.sys.p"]]`%) and `r odds.logged[["can.usr"]]` by users (`r odds.logged[["can.usr.p"]]`%). Note that the user cancellation rates between the not selected, random and inherited trips differed greatly between strata. The randomly selected trips in the *`r odds.can.usr.prop$Strata`* stratum were cancelled by users at a rate `r round(odds.can.usr.prop$r_n_prop, 1)`$\times$ higher than trips that were randomly not selected. 
\
The number of completed trips that were randomly selected for monitoring or from inherited monitoring, as well as the number of trips that were waived, are summarized in `r tbl_nums('logged.table', display = "cite")`. It is notable that `r logged.table %>% filter(Strata == "OB FIXED GOA") %>% pull(-2) %>% unique()`% of monitored trips in the *OB FIXED GOA* stratum and `r logged.table %>% filter(Strata == "OB TRW GOA") %>% pull(-2) %>% unique()`% of monitored trips in the *OB TRW GOA* stratum were inherited. This is in contrast to the *EM FIXED* strata where fewer than `r logged.table %>% filter(Strata %like% "EM FIXED") %>% pull(-2) %>% max() %>% ceiling()`% of monitored trips were inherited. The monitoring inheritance system is necessary to compensate for trip cancellations to achieve the monitoring rates prescribed by the ADP. However, the monitoring program was designed with the intention of employing a randomized sampling design, and trips with inherited monitoring are not selected randomly. \
\
The extent to which trip-selections are changed from the time they are entered can be determined by comparing the rate of trip monitoring expected from (1) random selection of all logged trips (initial random selection) and (2) random selection of remaining trips after cancellations, inherits, and waivers. In any case, the proportion of trips selected to be monitored should fall within what would be expected given the binomial distribution (since each trip is either selected or not selected). The rates obtained (%, with associated *p*-value based on the binomial distribution) in the initial selection process by ODDS were within expected ranges for all strata (`r tbl_nums('Selection.table', display = "cite")`). Moreover, all strata achieved final rates within the expected ranges after waivers. It is worth noting that the *OB FIXED BSAI* stratum had an elevated initial selection rate by ODDS and a high number of trips with inherited monitoring. The *OB FIXED GOA* stratum had a large number of cancellations that was counteracted by a large number of trips with inherited monitoring. A time series of ODDS initial selection rates and final realized rates is presented in `r fig_nums('odds.by.date.plot', display = "cite")`.

## Evaluation of Coverage Rates  
```{r Coverage.tbl, message=FALSE, warning=FALSE, include=FALSE}
# Coverage table has number of total vessels (V), sampled vessels (v), total trips (N), sampled trips (n), and percent observed by deployment method (trip vs. vessel-selection):

#' \TODO *2024 AR* `Fix EM Trawl GOA calculations - VALHALLA currently not handling tenders well`
#' For now using observed assignment based on observer data (column: OBS_SALMON_CNT_FLAG)
#' This means when calculating summaries, we need to remove EM trawl and calculate it on it's
#' own and then join back to the dataframe

#' * 2024 AR * `Temporary fix until VALHALLA trip logic is updated`
# Make delivery data look like trip data
# Assuming one landing report per delivery
# Delivery = when fish actually hit the processing plant (e.g., when a tender delivers all CV offloads)
# Will still summarize V and v by CV, but N and n will be by tender (when applicable) and CV
work.offload.trw_em <- work.offload %>%
  # First, use REPORT_ID to replace TRIP_ID
  mutate(TRIP_ID = as.character(REPORT_ID),
         VESSEL_ID = as.integer(CV_ID),
         COVERAGE_TYPE = if_else(STRATA == "EM TRW GOA (EFP)", "PARTIAL", "FULL")) %>%
  # Replace TRIP_ID with T_REPORT_ID if present to group up shoreside tender deliveries
  mutate(TRIP_ID = ifelse(!is.na(T_REPORT_ID), T_REPORT_ID, TRIP_ID)) %>%
  select(COVERAGE_TYPE, STRATA, TRIP_ID, VESSEL_ID,
         OBSERVED_FLAG = OBS_SALMON_CNT_FLAG,
         TENDER = OFFLOAD_TO_TENDER_FLAG,
         FMP, PORT_CODE, REPORT_ID)
#' Make a joining table for the analyses
work.data.tem.join <- unique(work.offload.trw_em[, .(TRIP_ID, REPORT_ID, OBSERVED_FLAG)])


# Add back to work.data
work.data.offload <- work.data %>%
  filter(!grepl("EM TRW", STRATA)) %>%
  mutate(TRIP_ID = as.character(TRIP_ID)) %>%
  rbind(work.offload.trw_em, fill = TRUE)

#' *FULL COVERAGE*
Coverage.Table.full <- bind_rows(
  # Coverage rates for vessels in the full coverage category (both regulatory
  # and voluntary full coverage)
  work.data.offload %>%
    filter(COVERAGE_TYPE == "FULL") %>% 
    group_by(COVERAGE_TYPE, STRATA) %>%
    summarize(V = length(unique(VESSEL_ID)),
              v = length(unique(VESSEL_ID[OBSERVED_FLAG == "Y"])),
              N = length(unique(TRIP_ID)),
              n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"]))) %>%
    mutate(Pct.obs = (n/N) * 100,
           deployment.method = "Trip-Selection",
           Rate = 1,
           meets.expected = ifelse(Pct.obs < 100, "No - lower than expected", "Yes")) %>%
    select(COVERAGE_TYPE, STRATA, deployment.method, 
           V, v, N, n, Pct.obs, Rate, meets.expected),
  # Totals for vessels in the full coverage category
  work.data.offload %>%
    filter(COVERAGE_TYPE == "FULL") %>%
    summarize(V = length(unique(VESSEL_ID)),
              v = length(unique(VESSEL_ID[OBSERVED_FLAG == "Y"])),
              N = length(unique(TRIP_ID)),
              n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"]))) %>%
    mutate(COVERAGE_TYPE = "Full coverage total",
           Pct.obs = (n/N) * 100,
           deployment.method = "Trip-Selection")
) %>%
  arrange(desc(STRATA)) %>%
  ungroup()

#' *PARTIAL COVERAGE TRIP-SELECTION*
Coverage.Table.partial <- bind_rows(
  # By strata for vessels in the partial coverage category (trip-selection)
  work.data.offload %>% 
    filter(COVERAGE_TYPE == "PARTIAL" & STRATA %in% partial$STRATA) %>%
    left_join(partial %>% select(STRATA, Rate) %>% distinct(),
              by = join_by(STRATA)) %>%
    group_by(COVERAGE_TYPE, STRATA) %>%
    summarize(V = length(unique(VESSEL_ID)),
              v = length(unique(VESSEL_ID[OBSERVED_FLAG == "Y"])),
              N = length(unique(TRIP_ID)),
              n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"])),
              Rate = unique(Rate)) %>%
    mutate(Pct.obs = (n/N) * 100,
           deployment.method = "Trip-Selection") %>% 
    # Add binomial distribution upper and lower 95th percentiles for each partial coverage stratum
    group_by(COVERAGE_TYPE, STRATA) %>%
    mutate(
      min.expected = round(
        stats::binom.test(x = n, 
                          n = N, 
                          p = Rate, 
                          alternative = "two.sided")$conf.int[1], digits = 4) * 100,
      max.expected = round(
        stats::binom.test(x = n, 
                          n = N, 
                          p = Rate,
                          alternative = "two.sided")$conf.int[2], digits = 4) * 100,
      meets.expected = case_when(Rate * 100 > min.expected & Rate * 100 < max.expected ~ "Yes",
                                 Rate * 100 < min.expected ~ "No - higher than expected",
                                 Rate * 100 > max.expected ~ "No - lower than expected")) %>%
    select(COVERAGE_TYPE, STRATA, deployment.method, 
           V, v, N, n, Pct.obs, Rate, 
           min.expected, max.expected, meets.expected)
  ,
  # Totals for partial coverage category
  Coverage.Table.partial.total <- work.data.offload %>%
    filter(COVERAGE_TYPE == "PARTIAL" & STRATA %in% partial$STRATA) %>%
    summarize(V = length(unique(VESSEL_ID)),
              v = length(unique(VESSEL_ID[OBSERVED_FLAG == "Y"])),
              N = length(unique(TRIP_ID)),
              n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"]))) %>%
    mutate(COVERAGE_TYPE = "Partial coverage total",
           Pct.obs = (n/N) * 100,
           deployment.method = "Trip-Selection") 
) %>%
  arrange(match(STRATA, strata.order)) %>%
  ungroup()

#' *ZERO COVERAGE* 
Coverage.Table.zero <-
  work.data.offload %>%
    filter(STRATA %in% c("ZERO", "Zero EM Research")) %>% #'
    mutate(STRATA = as.character(STRATA)) %>% 
    mutate(STRATA = ifelse(STRATA == "ZERO", "Zero coverage", STRATA)) %>% 
    group_by(COVERAGE_TYPE, STRATA) %>%
    summarize(V = length(unique(VESSEL_ID)),
              v = length(unique(VESSEL_ID[OBSERVED_FLAG == "Y"])),
              N = length(unique(TRIP_ID)),
              n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"]))) %>%
    mutate(Pct.obs = (n/N) * 100,
           deployment.method = "Trip-Selection",
           Rate = 0,
           meets.expected = ifelse(Pct.obs > 0, "No - higher than expected", "Yes")) %>%
    select(COVERAGE_TYPE, STRATA, deployment.method, 
           V, v, N, n, Pct.obs, Rate, meets.expected) %>%
  ungroup()

# Totals with EM (number of vessels is not additive)
Coverage.Table.TOTALS <- work.data.offload %>%
  summarize(V = length(unique(VESSEL_ID)),
            v = length(unique(VESSEL_ID[OBSERVED_FLAG == "Y"])),
            N = length(unique(TRIP_ID)),
            n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"]))
  ) %>%
  mutate(COVERAGE_TYPE = "Total",
         STRATA = "Total",
         Pct.obs.trips = format((n/N) * 100, digits = 1, nsmall = 2),
         Pct.obs.vessels = format((v/V) * 100, digits = 1, nsmall = 2),
         Pct.obs = paste0(Pct.obs.trips, "% trips/deliveries; ", Pct.obs.vessels, "% vessels")) %>%
  ungroup()

#' *Combine all the coverage tables*
Coverage.Table <- bind_rows(Coverage.Table.full, 
                            Coverage.Table.partial,
                            Coverage.Table.zero,
                            ) %>%
  mutate(Pct.obs = format(Pct.obs, digits = 1, nsmall = 2)) %>%
  bind_rows(Coverage.Table.TOTALS) %>%
  # Formatting (change things to characters so they knit to Word without weird
  # formatting issues)
  mutate(V = as.character(V),
         v = as.character(v),
         N = as.character(N),
         n = as.character(n),
         #' *FLAG* binom.test method
         #Rate = ifelse(is.na(Rate), NA, format(Rate * 100, digits = 1, nsmall = 1)),
         #' *FLAG* qbinom method
         Rate = ifelse(is.na(Rate), NA, 
                       ifelse(deployment.method == "Trip-Selection", 
                              format(Rate * 100, digits = 1, nsmall = 2),
                              format(Rate, digits = 0, nsmall = 0))),
         min.expected = ifelse(is.na(min.expected), NA, 
                               ifelse(deployment.method == "Trip-Selection", 
                                      format(min.expected, digits = 1, nsmall = 2),
                                      format(min.expected, digits = 0, nsmall = 0))),
         max.expected = ifelse(is.na(max.expected), NA, 
                               ifelse(deployment.method == "Trip-Selection", 
                                      format(max.expected, digits = 1, nsmall = 2),
                                      format(max.expected, digits = 0, nsmall = 0))),
         STRATA = case_when(STRATA == "FULL" ~ "Full", 
                            STRATA == "EM TRW GOA (EFP)" ~ "EM TRW GOA (EFP)*",
                            STRATA == "EM TRW BSAI (EFP)" ~ "EM TRW BSAI (EFP)*", 
                            TRUE ~ STRATA),
         COVERAGE_TYPE = case_when(STRATA == "Zero coverage" ~ "Zero coverage",
                                   COVERAGE_TYPE == "FULL" ~ "Full coverage",
                                   COVERAGE_TYPE == "PARTIAL" ~ "Partial coverage",
                                   TRUE ~ COVERAGE_TYPE)) %>%
  mutate(STRATA = as.factor(STRATA),
         COVERAGE_TYPE = as.factor(COVERAGE_TYPE)) %>%
  select(`Coverage` = COVERAGE_TYPE, 
         Strata = STRATA,
         V, v, N, n,
         `Expected` = Rate,
         `Realized` = Pct.obs,
         `Lower limit`= min.expected,
         `Upper limit` = max.expected, 
         `Realized meets expected?` = meets.expected)

# Create an empty factor level so that you can get rid of NAs in the table.
Coverage.Table$Coverage = factor(Coverage.Table$Coverage, levels = c(levels(Coverage.Table$Coverage), " "))
Coverage.Table$Strata = factor(Coverage.Table$Strata, levels = c(levels(Coverage.Table$Strata), " "))
Coverage.Table <- replace(Coverage.Table, is.na(Coverage.Table), " ")

rm(Coverage.Table.full, Coverage.Table.partial, Coverage.Table.partial.total, Coverage.Table.TOTALS, Coverage.Table.zero)

# Check full coverage to see why < 100%
full.cover.no.obs <- work.data %>%
  filter(COVERAGE_TYPE == "FULL" & OBSERVED_FLAG == "N") %>%
  select(TRIP_ID, VESSEL_ID, PERMIT, AGENCY_GEAR_CODE, FMP, MANAGEMENT_PROGRAM_CODE, 
         PROCESSING_SECTOR, TRIP_TARGET_CODE, LENGTH_OVERALL) %>%
  distinct()

```

```{r coverage_paragraph, message=FALSE, warning=FALSE, include=FALSE}
# Objects referenced in-text:

# Coverage trip and vessel totals (including EM)
totl_obs_trips <- Coverage.Table %>% filter(Strata == "Total" & Coverage == "Total") %>% select(n) %>% as.numeric() %>%
  prettyNum(big.mark = ",")

perc_trips_obs <- Coverage.Table %>% 
  filter(Strata == "Total" & Coverage == "Total") %>% 
  transmute(percent.obs = format(as.numeric(n)/as.numeric(N) * 100, digits = 1, nsmall = 1)) %>% 
  as.numeric()

totl_obs_vessels <- Coverage.Table %>% filter(Strata == "Total" & Coverage == "Total") %>% select(v) %>% as.numeric() %>%
  prettyNum(big.mark = ",")

perc_vessels_obs <- Coverage.Table %>% 
  filter(Strata == "Total" & Coverage == "Total") %>%
  transmute(percent.obs = format(as.numeric(v)/as.numeric(V) * 100, digits = 1, nsmall = 1)) %>% 
  as.numeric()
```

```{r data_timeliness, message=FALSE, warning=FALSE, include=FALSE, eval=TRUE}
data_timeliness.strata_order <- c(
  "OB FIXED BSAI", "OB FIXED GOA", "OB TRW BSAI", "OB TRW GOA", 
  "EM FIXED BSAI", "EM FIXED GOA", "EM TRW BSAI (EFP)", "EM TRW GOA (EFP)"
)

# Set the ordering of the strata and calculate summary statistics
data_timeliness <- data_timeliness %>% 
  mutate(STRATA = factor(STRATA, levels = data_timeliness.strata_order))

times_smry <- data_timeliness %>%
              group_by(YEAR, COVERAGE_TYPE, STRATA) %>%
              summarise(median = round(median(data_timeliness), 2), 
                        mean = round(mean(data_timeliness), 2), 
                        .groups = "drop") %>%
  mutate(STRATA = factor(STRATA, levels = data_timeliness.strata_order))

times_figure <- ggplot(data_timeliness, aes(x = data_timeliness)) +
                facet_wrap(STRATA ~ ., scales = "free_y", ncol = 2) +
                geom_histogram(bins = 20, fill = "gray70") +
                geom_vline(data = times_smry, aes(xintercept = median), color = "black") +
                geom_vline(data = times_smry, aes(xintercept = mean), color = "red", lty = 2) +
                geom_text(data = times_smry,
                          aes(x = median, y = Inf, label = paste(format(median, nsmall = 2))), 
                          angle = 90, hjust = 1.2, vjust = -0.3) +
                geom_text(data = times_smry, 
                          aes(x = mean, y = Inf, label = paste(format(mean, nsmall = 2))),
                          angle = 90, hjust = 1.2, vjust = 1.1) +
                labs(x = "Data timeliness (days)", y = "Count") +
                theme_bw()

```

```{r review_timeliness, message=FALSE, warning=FALSE, include=FALSE, eval=TRUE}
review_times_smry <- data_timeliness %>%
              filter(STRATA %like% c("EM FIXED")) %>% 
              group_by(YEAR, COVERAGE_TYPE, STRATA) %>%
              summarise(median = round(median(review_timeliness), 2), 
                        mean = round(mean(review_timeliness), 2), 
                        .groups = "drop") %>%
  mutate(STRATA = factor(STRATA, levels = data_timeliness.strata_order))

review_times_figure <- ggplot(filter(data_timeliness, STRATA %like% c("EM FIXED")), aes(x = review_timeliness)) +
                facet_wrap(STRATA ~ ., scales = "free_y", ncol = 2) +
                geom_histogram(bins = 20, fill = "gray70") +
                geom_vline(data = review_times_smry, aes(xintercept = median), color = "black") +
                geom_vline(data = review_times_smry, aes(xintercept = mean), color = "red", lty = 2) +
                geom_text(data = review_times_smry,
                          aes(x = median, y = Inf, label = paste(format(median, nsmall = 2))), 
                          angle = 90, hjust = 1.2, vjust = -0.3) +
                geom_text(data = review_times_smry, 
                          aes(x = mean, y = Inf, label = paste(format(mean, nsmall = 2))),
                          angle = 90, hjust = 1.2, vjust = 1.1) +
                labs(x = "Review timeliness (days)", y = "Count") +
                theme_bw()

```
This section compares the coverage rate achieved against the expected coverage rates. Data used in this evaluation are stored within the Catch Accounting System (CAS; managed by the AKRO), the Observer Program database (NORPAC; managed by the AFSC), and eLandings (under joint management by the Alaska Department of Fish and Game, the International Pacific Halibut Commission and NMFS). Separate rate evaluations are conducted depending on whether the unit of observer deployment was at-sea fishing trips or dockside deliveries of pollock.\
\
In this document, trips in the *EM FIXED* strata are considered successfully monitored if at least some video was reviewed. The rationale for defining monitored trips this way is that it is most similar to the way in which trips in other strata are considered observed (i.e., irrespective of whether or not haul information or usable species composition data were collected). Deliveries in the *EM TRW (EFP)* strata were considered successfully monitored if an observer was able to collect salmon information from a delivery.\
\
In combination across all strata, coverage levels, and fishery monitoring tools, `r totl_obs_trips` trips/deliveries (`r perc_trips_obs`%) and `r totl_obs_vessels` vessels (`r perc_vessels_obs`%) were successfully monitored among all fishing in the Federal fisheries off Alaska in `r year` (`r tbl_nums('Coverage.Table', display = "cite")`). This compares to a total of 3,780 trips/deliveries (43.7%) and 463 vessels (50.2%) monitored in `r year - 1` (AFSC and AKRO `r year`).\
\
In `r year`, the Observer Program had `r total.strat` different deployment strata to evaluate (`r tbl_nums('Coverage.Table', display = "cite")`). There were two *Full coverage* strata; vessels that were required to have full coverage (e.g., AFA vessels and most catcher/processors) and BSAI trawl catcher vessels that opted into full coverage (i.e., *EM TRW BSAI [EFP]* under regulations *50 CFR 679.51(a)(4)*). There were eight partial coverage strata: four observed strata, three EM strata, defined by gear designation, and also one *Zero coverage* stratum that included jig vessels and vessels under 40 ft LOA that are not monitored.\
\
Evaluations for the full coverage category and zero-selection pool are straightforward --- either the coverage achieved was equal to 100% or 0%, respectively, or it was not. The program achieved `r Coverage.Table %>% filter(Strata == "Full") %>% pull(Realized) %>% trimws()`% coverage in its *Full coverage* stratum (`r tbl_nums('Coverage.Table', display = "cite")`). One trip was not monitored in the *Full coverage* stratum (see Appendix C). The program achieved compliance with the *Zero coverage* stratum (`r tbl_nums('Coverage.Table', display = "cite")`).\
\
For the partial coverage strata, under the assumption that the deployment was randomized, a 95% confidence interval computed from the realized coverage rates (assuming a binomial distribution for monitored trips) will contain the actual deployment rate 95% percent of the time. If expected coverage levels were within the 95% confidence intervals, then we conclude that realized coverage rates did not deviate from expected coverage rates.  Trips in the *EM FIXED* strata were considered covered only if the catch data from monitored trips were reviewed by the Pacific States Marine Fisheries Commission (PSMFC).\
\
Coverage rates were consistent with expected values in six of the seven partial coverage strata that are subject to monitoring, indicating that the deployment of monitoring as intended by the `r year` ADP was largely successful. Coverage rates were lower than expected in the *EM FIXED BSAI* stratum (`r tbl_nums('Coverage.Table', display = "cite")`) due to a temporary understaffing of EM video reviewers.\
\
Unlike observed trips, the coverage rate for EM is based on information provided from the PSFMC that is available to analysts in CAS at the time of writing. In 2024, the median time between receipt and completion of review was `r times_smry %>% filter(STRATA == "EM FIXED BSAI") %>% pull(median)` days for *EM FIXED BSAI* and `r times_smry %>% filter(STRATA == "EM FIXED GOA") %>% pull(median)` days for *EM FIXED GOA* `r fig_nums('times_figure', display = "cite")`. The mean time between receipt and completion of review was `r times_smry %>% filter(STRATA == "EM FIXED BSAI") %>% pull(mean) %>% round()` days for *EM FIXED BSAI* and `r times_smry %>% filter(STRATA == "EM FIXED GOA") %>% pull(mean) %>% round()` days for *EM FIXED GOA*.\
\
Data timeliness values for the 2024 *EM FIXED* strata cannot be directly compared to data timeliness values for these strata in prior years, in part because 2024 was the first year in which the *EM FIXED BSAI* and *EM FIXED GOA* strata were used (in contrast to the *EM HAL* and *EM POT* strata in prior years). More importantly, data timeliness values for these strata in 2024 should not be compared to prior years because, in the process of compiling this report, we discovered that fixed-gear EM data were missing from CAS. Upon further investigation, we discovered that a coding error was preventing those data from being transferred from the AFSC to CAS. Therefore, we have included `r fig_nums('review_times_figure', display = "cite")`, which shows review timeliness for the EM FIXED strata. Review timeliness differs from data timeliness in that it describes the time between a trip ending and data being reviewed by PSMFC staff (in contrast to data timeliness, which describes the time between a trip ending and data being available to CAS). In prior years, review timeliness and data timeliness were approximately equal. However, the coding error discovered in the process of compiling this report resulted in much data being received long after the associated trips had been reviewed by PSMFC staff. This explains why review timeliness for these strata is much better than data timeliness. Median and mean review timeliness for *EM FIXED BSAI* was `r review_times_smry %>% filter(STRATA == "EM FIXED BSAI") %>% pull(median)` days and `r review_times_smry %>% filter(STRATA == "EM FIXED BSAI") %>% pull(mean) %>% round()` days, respectively (`r fig_nums('review_times_figure', display = "cite")`). Median and mean review timeliness for *EM FIXED GOA* was `r review_times_smry %>% filter(STRATA == "EM FIXED GOA") %>% pull(median)` days and `r review_times_smry %>% filter(STRATA == "EM FIXED GOA") %>% pull(mean) %>% round()` days, respectively. Although the difference between data timeliness and review timeliness was not that large for *EM FIXED BSAI*, most fixed-gear EM trips take place in the *EM FIXED GOA* stratum. Staff at AFSC have remedied the coding error responsible for this delay, so we do not expect a similar delay for fixed- gear EM data in 2025.\
\
Although review timeliness for the *EM FIXED strata* cannot be directly compared to data timeliness for these strata in prior years, the terms are similar enough that some conclusions can be made. It appears that review timeliness for these strata in 2024 was much better than data timeliness in prior years. We attribute this result to the fact that in 2024, NMFS directed PSMFC to prioritize the review of current-year fixed-gear EM trips, rather than attempting to finish review of the prior year before beginning on the current year. PSMFC has carried this prioritization forward into 2025, and, with the resolution of the coding error identified in the process of compiling this report, we expect data timeliness for the *EM FIXED* strata to be better in 2025 than it was in 2024.\
\
Note that there are several reasons why the total number of trips and the final monitoring rates presented in this section differ with what was presented earlier within ODDS. As previously mentioned, trips fishing CDQ/IFQ in multiple NMFS areas are required to be monitored and are not randomly selected, thus they were excluded from the evaluation. Additionally, there is no robust link between the ODDS database and eLandings, therefore, trips in ODDS cannot be linked directly to realized landings in eLandings, which inform the trip identification numbers created by CAS. The FMSC has recommended in past reports that this linkage be established. Moreover, ODDS trips are sometimes not logged as required (see Chapter 5, section 1.5.2, Table 5-3) and records are not created in ODDS after-the-fact. Finally, if *EM FIXED* strata trips are monitored but not reviewed, discrepancies in the realized selection rates and coverage rates will occur. 

### Coverage Rates for Dockside Monitoring
```{r pollock.observed.tbl, message=FALSE, warning=FALSE, include=FALSE}
# Get all pollock landings and label observed status  
# Note: tender is intentionally excluded
Pollock.landing.data <- work.data %>%
  
  # Check for non-pollock EM trawl offloads
  {
    filtered_data <- work.data %>%
      filter(grepl("EM TRW", STRATA) & !(TRIP_TARGET_CODE %in% c("B", "P"))) %>%
      distinct(REPORT_ID)

    non_pollock_count <- nrow(filtered_data)

    if (non_pollock_count > 0) {
      cat("\033[33m", paste(non_pollock_count, "EM trawl offload(s) non-pollock\033[0m")) # yellow
    } else {
      cat("\033[32mAll EM trawl deliveries are pollock\033[39m\n")
    }
  .} %>%
  
  # Observer Pollock Trawl Trips
  filter(STRATA %in% c("OB TRW BSAI", "OB TRW GOA", "FULL") &
           SOURCE_TABLE == "Y" & # Retained catch
           AGENCY_GEAR_CODE %in% c("PTR", "NPT") &
           MANAGEMENT_PROGRAM_CODE %in% c("AFA", "OA") &
           PROCESSING_SECTOR == "S" &
           TRIP_TARGET_CODE %in% c("B", "P"),
           TENDER == "N") %>%
  distinct(REPORT_ID, COVERAGE_TYPE, STRATA, FMP, PORT_CODE, TENDER) %>%
  
  # Check for duplicates in REPORT_ID
  {
    if (any(duplicated(.$REPORT_ID))) {
      cat(" \033[31mDuplicate REPORT_IDs found\033[39m\n")
    } else {
      cat(" \033[32mNo duplicate REPORT_IDs\033[39m\n")
    }
  .} %>%
  
  # Combine observer pollock trips with EM TRW BSAI trips
  # EM trawl is a pollock fishery, but deliveries may not always be marked as B or P
  #'* 2024 AR* 1 cod offload in BSAI
  rbind(work.data %>%
          filter(STRATA == "EM TRW BSAI (EFP)") %>%
          distinct(REPORT_ID, COVERAGE_TYPE, STRATA, FMP, PORT_CODE, TENDER)) %>%
  
  # salmon.landings.obs is a list of REPORT_IDs for which dockside sampling for salmon occurred (all sectors)
  mutate(Observed.type = ifelse(REPORT_ID %in% salmon.landings.obs$REPORT_ID, "Observed", "Not_Observed")) %>%
  
  # Combine with EM TRW GOA trips
  rbind(
    work.offload.trw_em %>%
      filter(STRATA == "EM TRW GOA (EFP)") %>%
      select(REPORT_ID = TRIP_ID, COVERAGE_TYPE, STRATA, FMP, PORT_CODE, Observed.type = OBSERVED_FLAG, TENDER) %>%
      distinct() %>%
      mutate(Observed.type = if_else(Observed.type == "Y", "Observed", "Not_Observed"),
             COVERAGE_TYPE = case_when(
               grepl("EM TRW GOA", STRATA) & TENDER == "Y" & COVERAGE_TYPE == "PARTIAL" ~ "Partial (tender)",
               grepl("EM TRW GOA", STRATA) & TENDER == "N" & COVERAGE_TYPE == "PARTIAL" ~ "Partial (non-tender)",
               TRUE ~ COVERAGE_TYPE)),
    fill = TRUE) %>%
  
  # Label FMP port, and coverage type
  mutate(FMP = fct_recode(factor(FMP), `Bering Sea` = "BSAI", `Gulf of Alaska` = "GOA"),
         PORT_CODE = fct_recode(factor(PORT_CODE),
                                `Kodiak` = "KOD", 
                                `Dutch Harbor` = "DUT", 
                                `Akutan` = "AKU", 
                                `Sand Point` = "SPT", 
                                `King Cove` = "KCO",
                                `False Pass` = "FSP",
                                `Inshore floating processor` = "IFP"),
         COVERAGE_TYPE = fct_recode(factor(COVERAGE_TYPE), `Full` = "FULL", `Partial` = "PARTIAL"))

# Summarize offloads by FMP, coverage type, port, and observed status
Pollock.monitored <- rbind(
  # Stratum and Port
  Pollock.landing.data %>%
    group_by(FMP, COVERAGE_TYPE, STRATA, PORT_CODE) %>%
    summarize(N.pollock = n(),
              N.pollock.obs = sum(Observed.type == "Observed")) %>%
    mutate(pct.obs = format((N.pollock.obs / N.pollock) * 100, digits = 1, nsmall = 2)) %>% 
    select(FMP, COVERAGE_TYPE, STRATA, PORT_CODE, N.pollock, N.pollock.obs, pct.obs) %>%
    ungroup(),
  # Stratum-specific totals
  Pollock.landing.data %>%
    group_by(COVERAGE_TYPE, STRATA) %>%
    summarize(FMP = as.factor("Total"),
              PORT_CODE = as.factor(""),
              N.pollock = n(),
              N.pollock.obs = sum(Observed.type == "Observed")) %>%
    mutate(pct.obs = format((N.pollock.obs / N.pollock) * 100, digits = 1, nsmall = 2)) %>%
    select(FMP, COVERAGE_TYPE, STRATA, PORT_CODE, N.pollock, N.pollock.obs, pct.obs) %>% 
    ungroup(),
  
  Pollock.landing.data %>%
    filter(STRATA == "EM TRW GOA (EFP)") %>%
    summarize(FMP = as.factor("Total"),
              COVERAGE_TYPE = as.factor("Partial"),
              STRATA = as.character("EM TRW GOA (EFP)"),
              PORT_CODE = as.factor(""),
              N.pollock = n(),
              N.pollock.obs = sum(Observed.type == "Observed")) %>%
    mutate(pct.obs = format((N.pollock.obs / N.pollock) * 100, digits = 1, nsmall = 2)) %>%
    select(FMP, COVERAGE_TYPE, STRATA, PORT_CODE, N.pollock, N.pollock.obs, pct.obs) %>% 
    ungroup())

# Isolate observed strata (Non-EM, both full and partial coverage)

Pollock.observed <- rbind(
  filter(Pollock.monitored, COVERAGE_TYPE == "Full" & !grepl("EM TRW", STRATA)),
  filter(Pollock.monitored, COVERAGE_TYPE == "Partial" & !grepl("EM TRW", STRATA)))

# Rename columns and format
Pollock.observed.table <- Pollock.observed %>%
  mutate(N.pollock = as.character(N.pollock),
         N.pollock.obs = as.character(N.pollock.obs),
         FMP = case_when(FMP == "Bering Sea" ~ "BSAI",
                         COVERAGE_TYPE == "Full" ~ "BSAI total",
                         FMP == "Gulf of Alaska" ~ "GOA",
                         COVERAGE_TYPE == "Partial" ~ "GOA total",
                         TRUE ~ FMP)) %>% 
  mutate(COVERAGE_TYPE = if_else(FMP %like% "total", "", COVERAGE_TYPE)) %>%
  select(`FMP` = FMP, 
         `Coverage category` = COVERAGE_TYPE,
         `Port` = PORT_CODE, 
         `Total deliveries (*N*)` = N.pollock,
         `Observed deliveries (*n*)` = N.pollock.obs,
         `% observed` = pct.obs)

# Isolate EFP strata
Pollock.efp <- rbind(
  filter(Pollock.monitored, COVERAGE_TYPE == "Full" & grepl("EM TRW", STRATA)),
  filter(Pollock.monitored, COVERAGE_TYPE == "Partial (non-tender)" & grepl("EM TRW", STRATA)),
  filter(Pollock.monitored, COVERAGE_TYPE == "Partial (tender)" & grepl("EM TRW", STRATA)),
  filter(Pollock.monitored, COVERAGE_TYPE == "Partial" & grepl("EM TRW", STRATA)))

# Rename columns and format
Pollock.efp.table <- Pollock.efp %>% 
  mutate(N.pollock = as.character(N.pollock),
         N.pollock.obs = as.character(N.pollock.obs),
         FMP = case_when(FMP == "Bering Sea" ~ "BSAI",
                         COVERAGE_TYPE == "Full" ~ "BSAI total",
                         FMP == "Gulf of Alaska" ~ "GOA",
                         COVERAGE_TYPE == "Partial (non-tender)" ~ "GOA (non-tender) total",
                         COVERAGE_TYPE == "Partial (tender)" ~ "GOA (tender) total",
                         COVERAGE_TYPE == "Partial" ~ "GOA total",
                         TRUE ~ FMP)) %>% 
  mutate(COVERAGE_TYPE = if_else(FMP %like% "total", "", COVERAGE_TYPE)) %>%
  select(`FMP` = FMP, 
         `Coverage category` = COVERAGE_TYPE,
         `Port` = PORT_CODE, 
         `Total deliveries (*N*)` = N.pollock,
         `Observed deliveries (*n*)` = N.pollock.obs,
         `% observed` = pct.obs)
```

For this analysis, pollock deliveries in the *OBS TRW* strata were defined as any delivery where the predominant species was pollock in eLandings, catch was processed at a shoreside plant, and the management program was AFA (for the BSAI) or Open Access (for the GOA). Pollock deliveries in the *EM TRW (EFP)* strata were defined as any delivery where the vessel was fishing under the Trawl EM EFP, including deliveries to a tender vessel that subsequently offloaded at a shoreside plant. In `r year`, 100% of full coverage pollock deliveries were monitored for salmon in both strata, meeting expectations (`r tbl_nums('Pollock.observed.table', display = "cite")`, `r tbl_nums('Pollock.efp.table', display = "cite")`).\
\
Evaluations of the partial coverage category for dockside monitoring are not as straightforward as for full or zero coverage. As a matter of policy, no tender deliveries were monitored in the *OBS TRW* strata and these deliveries are not included in this chapter. However, tender deliveries in the *EM TRW (EFP)* strata were monitored and are included in reporting. While it may seem intuitive that the expected coverage rate for non-tendered deliveries within the *OB TRW GOA* strata should be equal to the programmed trip selection rate of `r strata.rates$Rate[4] * 100`%, this assumption is not necessarily true because observers are not deployed specifically into the pollock fishery, but into the entire trawl fishery, and the relationship between the number of deliveries and trips is not expected to be constant, especially when measured across ports. Therefore, we present the dockside monitoring rates for non-tendered *OB TRW GOA* pollock deliveries but make no comparison to the expected deployment rates (`r tbl_nums('Pollock.observed.table', display = "cite")`). For *EM TRW GOA (EFP)* deliveries, we present dockside monitoring rates for both tendered and non-tendered pollock deliveries which can be compared to expected monitoring rates for shoreside deliveries in the *EM TRW GOA (EFP)* strata (`r tbl_nums('Pollock.efp.table', display = "cite")`). In `r year`, `r sprintf("%.2f", as.numeric(Pollock.efp.table$'% observed'[13]))`% of *EM TRW GOA (EFP)* deliveries were monitored for salmon, meeting expectations. 

# Sample Quality
## Temporal Patterns in Trip-Selection  
```{r temporal_bias, message=FALSE, warning=FALSE, include=FALSE}
# Changed significantly for 2020 to incorporate multiple time periods  
# Some changes might be worth keeping though  
Max.ts.date <- work.data %>% 
  filter(STRATA %in% partial$STRATA &  STRATA != "EM TRW GOA (EFP)") %>%
  select(TRIP_ID, TRIP_TARGET_DATE, OBSERVED_FLAG, STRATA) %>%
  distinct() %>%
  group_by(TRIP_ID, OBSERVED_FLAG, STRATA) %>%
  mutate(MAX_TARGET_DATE = max(TRIP_TARGET_DATE)) %>%
  select(TRIP_ID, MAX_TARGET_DATE, OBSERVED_FLAG, STRATA) 

# Count the number of total and monitored trips that occurred at each date
For.ts.temporal.plot <- Max.ts.date  %>%
  group_by(STRATA, Date = MAX_TARGET_DATE) %>%
  summarise(N = length(unique(TRIP_ID)), # Number of total trips
            n = length(unique(TRIP_ID[OBSERVED_FLAG == "Y"]))) %>% # Number of observed trips 
  ungroup()

# Fill in missing dates using tidyr::complete()
For.ts.temporal.plot <- For.ts.temporal.plot %>% 
  mutate(Date = as.Date(Date)) %>% 
  complete(nesting(STRATA), Date = seq(as.Date(paste0(year, "-01-01")), as.Date(paste0(year, "-12-31")), by = "day")) %>%
  mutate(YEAR = year(Date))

# Fill in rates
For.ts.temporal.plot <- left_join(For.ts.temporal.plot,
                                  partial %>%
                                    select(c(YEAR, STRATA, Rate)) %>%
                                    distinct(),
                                  by = c("YEAR", "STRATA"))

# Count cumulative number of trips by stratum and year
For.ts.temporal.plot <- 
  For.ts.temporal.plot %>% 
  group_by(YEAR, STRATA) %>%
  mutate(N = ifelse(is.na(N), 0, N),
         n = ifelse(is.na(n), 0, n)) %>% 
  mutate(N.cuml = cumsum(N),
         n.cuml = cumsum(n)) %>% 
  ungroup()

# Calculate the expected number of trips  
# The number of trips that occured prior to the third time period
# is used as a vertical adjustment for the expectations on the plot  
For.ts.temporal.plot <- 
  For.ts.temporal.plot %>%
  group_by(YEAR) %>%
  mutate(n.expected = round((N.cuml * Rate), 0),
         Low95 = qbinom(0.025, N.cuml, Rate),
         High95 = qbinom(0.975, N.cuml, Rate),
         Outside.bounds = ifelse(n.cuml > High95 | n.cuml < Low95, 1, 0),
         Observed.days = ifelse(n > 0, 1, NA)) %>% 
  data.frame()

# Summarise days and portion of the year outside of expected
For.ts.temporal.summary <- 
  For.ts.temporal.plot %>% 
  group_by(YEAR, STRATA) %>% 
  reframe(N.outside = sum(Outside.bounds, na.rm = TRUE),
          Days = sum(Date >= as.Date(paste0(year, "-01-01"))))  %>%
  distinct() %>% 
  mutate(`% Outside` = round(N.outside / Days * 100, 1))

# Calculate and store the p.value from a binomial
for.binom <- 
  For.ts.temporal.plot %>% 
  group_by(YEAR, STRATA, Rate) %>% 
  summarize(n = sum(n), 
            N = sum(N)) %>% 
  mutate(p = format(round(stats::binom.test(n, N, Rate)$p.value[1], 3), nsmall = 3),
         estimate = format(round(stats::binom.test(n, N, Rate)$estimate[1], 3), digits = 3, nsmall = 3),
         # Results text with probability of success (estimate) & p-value
         text = paste0("*", STRATA, "* stratum expected rate = ", round(Rate, 3), 
                       ", realized rate = ", estimate, ", *p*-value = ", p)) %>% 
  ungroup()

# Interpretive text indicating whether the strata "were" or "were not" within
# expectations of the binomial distribution
passfail <- ifelse(any(for.binom$p < 0.05), "were not", "were")

```

```{r cum.temporal.plot, message=FALSE, warning=FALSE, include=FALSE}
# Add p-value to data.frame
For.ts.temporal.plot <- left_join(For.ts.temporal.plot,
                                  select(for.binom, c(YEAR, STRATA, p)), by = c("YEAR", "STRATA"))

# Order strata for plot
For.ts.temporal.plot <- transform(For.ts.temporal.plot,
                                  STRATA = factor(STRATA,levels = c("EM FIXED BSAI", "EM FIXED GOA", 
                                                                    "OB FIXED BSAI", "OB FIXED GOA",
                                                                    "OB TRW BSAI", "OB TRW GOA")))

# Add column for dates where coverage is outside expectation, and format p-values
For.ts.temporal.plot <- For.ts.temporal.plot %>%
                        mutate(Date.outside = if_else(Outside.bounds == 1, Date, as.Date(NA))) %>% 
                        mutate(p = ifelse(as.numeric(p) < 0.05, paste0(p, "*"), p)) %>%
                        mutate(p = ifelse(p == "0.000*", paste0("italic(p) < '0.001*'"), paste0("italic(p) == '", p, "'")))

# Create "dummy" dataframe to set limits for x axes
Temporal.plot.x_range <- data.frame(YEAR = rep(year + -1:0, each = 2)) %>%
  mutate(Date = as.Date(paste0(YEAR, c("-01-01", "-12-31"))))

# Single-Year Plot

Temporal.plot <- 
  ggplot(filter(For.ts.temporal.plot, YEAR == year), aes(x = Date)) +
  # Add lines for coverage rates
  geom_line(aes(y = n.cuml), col = "black", linewidth = 2) +
  geom_line(aes(y = n.expected), col = "gray") +
  geom_ribbon(aes(ymin = Low95, ymax = High95), alpha = 0.2) +
  geom_rug(aes(x = Date.outside), sides = "b") +
  facet_grid(factor(STRATA, levels = strata.order) ~., scales = "free", space = "free_x") +
    # Annotate p-values
  geom_text(aes(x = as.Date(-Inf), y = Inf, vjust = 1.25, label = p), 
            hjust = -0.1, size = 4, parse = TRUE, check_overlap = TRUE) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Cumulative trips in trip-selection strata\n") +
  xlab("\nDate")

```

The cumulative number of fishing trips in each stratum was multiplied by the stratum-specific selection rate to obtain the expected number of observed trips. Under the assumption that there is no temporal bias in observer coverage, 2.5% of values should be larger than the upper 95% confidence limit and 2.5% should be smaller than the lower limit. At the end of `r year` the number of observed trips was outside of this expected range in one of the seven partial coverage strata: *EM FIXED BSAI* (expected rate =  `r `trimws(Coverage.Table[8, 7])`%, realized rate = `r trimws(Coverage.Table[8, 8])`%; `r tbl_nums('Coverage.Table', display = "cite")` and `r fig_nums('Temporal.plot', display = "cite")`). The *EM FIXED BSAI* stratum was below the expected range for a brief period in July and then for the remainder of the year beginning in August. Coverage rates were within their expected ranges for 100% of the year for most other strata, with the exception of *OB FIXED GOA* and *EM FIXED GOA* which were very briefly above the expected range early in the year.

## Spatial Patterns in Trip-Selection  
```{r spatiotemporal_analysis, message=FALSE, warning=FALSE, include=FALSE, eval=TRUE}
# Spatial analyses, moved from spatiotemp_analysis.R for the 2024 AR

#' \TODO Need to replace 'interspersion' with 'proximity index' in this script and functions/spatiotemp_functions.R

## Setup for spatial and spatiotemporal analyses -----------------------------------------------------------------------

# Wrangle the Valhalla data set for spatiotemporal analyses
# First, re-assign TRIP_ID for the trawl EM strata to properly identify the sample unit as the delivery
work.data.st <- copy(work.data) 
work.data.st |>
  # Merge in new TRIP_ID and OBSERVED_FLAG using REPORT_ID
  _[, TRIP_ID_NEW := work.data.tem.join[work.data.st, TRIP_ID, on = .(REPORT_ID)]
  ][, OBS_FLAG_NEW := work.data.tem.join[work.data.st, OBSERVED_FLAG, on = .(REPORT_ID)]
  # Need TRIP_ID to stay numeric, so make a helper column
  ][, MAX_TRIP_ID := as.numeric(max(TRIP_ID))
  ][, TRIP_ID_NEW.int := MAX_TRIP_ID + .GRP, by = .(TRIP_ID_NEW)
  # For trawl EM trips only (those with TRIP_ID_NEW), replace TRIP_ID and OBSERVED_FLAG
  ][!is.na(TRIP_ID_NEW), ':=' (TRIP_ID = TRIP_ID_NEW.int, OBSERVED_FLAG = OBS_FLAG_NEW)
  # Remove helper columns
  ][, c("MAX_TRIP_ID", "TRIP_ID_NEW", "TRIP_ID_NEW.int", "OBS_FLAG_NEW") := NULL]
pc_effort_st <- spatiotemp_data_prep(work.data.st)
rm(work.data.st)

#' Realized Rates *Check that these match up with the realized rates in 4_AR_report*
realized_rates.val  <- pc_effort_st[, .(ADP, STRATA, TRIP_ID, OBSERVED_FLAG)] |>
  unique() |>
  _[, .(STRATA_N = .N, SAMPLE_RATE = sum(OBSERVED_FLAG == "Y")/.N), keyby = .(ADP, STRATA)]

# Create a subset table of realized monitored trips
realized_mon <- unique(pc_effort_st[OBSERVED_FLAG == "Y", .(ADP, STRATA, TRIP_ID)])

# Calculate the realized rates of each STRATUM x ADP
programmed_rates <- copy(data.table(partial)) |>
  # Remove gear type so STRATA and Rate are 1:1
  _[, -c("formatted_strat", "GEAR")] |>
  unique() |>
  _[, STRATA := gsub(" ", "_", STRATA)][]
setnames(programmed_rates, old = c("YEAR", "Rate"), new = c("ADP", "SAMPLE_RATE"))
# Add ZERO to programmed rates
programmed_rates <- rbind(programmed_rates, data.table(ADP = year, STRATA = "ZERO", SAMPLE_RATE = 0))
# Add STRATA_N to programmed_rates
programmed_rates[, STRATA_N := realized_rates.val[programmed_rates, STRATA_N, on = .(ADP, STRATA)]]
setcolorder(programmed_rates, c("ADP", "STRATA", "STRATA_N", "SAMPLE_RATE"))

#' Check to make sure all year/strata names match between rates objects
if(!fsetequal(unique(realized_rates.val[, .(STRATA, ADP)]), unique(programmed_rates[, .(STRATA, ADP)])) ){
  stop("STRATA and ADP in `realized_rates.val` does not match those in `programmed_rates!`")
}
#' Check to make sure all year/strata names match between pc_effort_st and rates objects!
if(!fsetequal(unique(pc_effort_st[, .(STRATA, ADP)]), unique(programmed_rates[, .(STRATA, ADP)])) ){
  stop("STRATA and ADP in `pc_effort_dt` does not match those in `programmed_rates!`")
}

# Specify ordering of strata. Adding a blank stratum for the sake of aligning BSAI/GOA in plots.
strata_levels <- c(
  "OB_FIXED_BSAI", "OB_FIXED_GOA", "OB_TRW_BSAI", "OB_TRW_GOA", "EM_FIXED_BSAI", "EM_FIXED_GOA", "", "EM_TRW_GOA_(EFP)"
)

#' Define boxes using the 2024 stratum definitions, post-stratifying the fixed gear strata by Gear type. Exclude the ZERO stratum as we don't need to simulate interspersion for a stratum with a 0% selection rate.

box_def <- define_boxes(
  data = pc_effort_st[STRATA != "ZERO"], space = c(2e5, 2e5), time = c("week", 1, "TRIP_TARGET_DATE", "LANDING_DATE"),
  year_col = "ADP", stratum_cols = "STRATA", geom = TRUE, ps_cols = "GEAR")

# Stratum-Specific Proximity ------------------------------------------------------------------------------------------

# Expected proximity given the programmed and realized rates
exp_interspersion.programmed <- calculate_interspersion(box_def, sample_rates = programmed_rates)
exp_interspersion.realized   <- calculate_interspersion(box_def, sample_rates = realized_rates.val)

# Actually realized proximity
real_interspersion.stratum   <- calculate_interspersion(box_def, realized_mon = realized_mon)

# Simulate trip selection to create distributions of proximity indices. This should take ~2 minutes to run.
# Programmed rates
sim.programmed.stratum <- calculate_interspersion(
  box_def, sample_rates = programmed_rates, sim_iter = sim_iter, seed = seed, hex_smry = TRUE)
# Realized rates
sim.realized.stratum <- calculate_interspersion(
  box_def, sample_rates = realized_rates.val, sim_iter = sim_iter, seed = seed, hex_smry = TRUE)

# Calculate density of both distributions
density.programmed.stratum <- calculate_density(sim.programmed.stratum, "dodgerblue", adjust = 2)
density.realized.stratum <- calculate_density(sim.realized.stratum, "green", adjust = 2)

# Combine distributions
density.stratum <- combine_distributions(density.realized.stratum, density.programmed.stratum)

# Plot distributions vs actually realized

#' \TODO Make better legend descriptions like density assuming realized rate, density assuming programmed rate, etc
plt.proximity.stratum <- plot_interspersion_density(
  density.stratum, real_interspersion.stratum, strata_levels) + 
  facet_nested_wrap(
    STRATA ~ ., dir = "h", ncol = 2, scales = "free", drop = F,
    labeller = labeller(STRATA = function(x) paste0(year, " : ", gsub("_", " ", x) )))
# Drop the uneeded facet (strip and panel)
# drop_facet(plt.proximity.stratum )
plt.proximity.stratum <- drop_facet(plt.proximity.stratum, drop = list(c(24,7), c(25, 7)))

# Calculate percentiles of the outcomes relative to both distributions

#' \TODO This table was not in the 2023 AR document. It can perhaps be used to annotate the plt.proximity.stratum figure as some of the distributions and/or expected values are very squished together, and it may be hard to tell whether outcomes are are in the tails or not.
#' \TODO Make the column names better, match what is in the document
percentile.stratum <- insp_percentile(
  real_interspersion.stratum[, -"ITER"], 
  sim.programmed.stratum, 
  sim.realized.stratum) |>
  _[, c("PROG_SIG", "REAL_SIG") := lapply(.SD, function(x) {
    fcase(
      x < 0.025, "<<", 
      x < 0.05, "<", 
      x > 0.975, ">>", 
      x > 0.95, ">", default = "")
    }), .SDcols = c("PERC_PROG", "PERC_REAL")
  # ecdf function can't have an open right side, so check if the realized rate is outside of 95% interval
  ][, PROG_SIG_CHECK := data.table::between(INSP, prog_0.025, prog_0.975)
  ][PROG_SIG_CHECK == T, PROG_SIG := ""
  ][, REAL_SIG_CHECK := data.table::between(INSP, real_0.025, real_0.975)
  ][REAL_SIG_CHECK == T, REAL_SIG := ""
  ][, c("prog_0.025", "prog_0.975", "real_0.025", "real_0.975", "PROG_SIG_CHECK", "REAL_SIG_CHECK") := NULL
  ][, STRATA := factor(STRATA, levels = strata_levels)][]
setorder(percentile.stratum, ADP, STRATA)
colnames(percentile.stratum) <- gsub("_", " ", colnames(percentile.stratum))
tbl.percentile.stratum <- percentile.stratum %>%
  flextable() %>%
  colformat_int(j = 1, big.mark = "") %>%
  colformat_double(j = 3, digits = 2) %>%
  colformat_double(j = c(5, 6, 7), digits = 4) %>%
  merge_v(j = 1) %>%
  hline(i = which(diff(percentile.stratum$ADP) > 0)) %>%
  fix_border_issues() %>% 
  autofit()

## Proximity Maps----

#' [NOTE] *These maps are for internal use to get more granularity of the patterns that led to the resulting proximity
proximity_maps <- plot_interspersion_map(
  box_def, real_interspersion.stratum, exp_interspersion.realized, exp_interspersion.programmed, ak_map.low_res.lst)
#' \TODO Add trip count text labels to $HEX_ID.realized plot cells 

# Save the plots as a multi-page PDF
if(F) {
  library(gridExtra)
  proximity_maps.box <- lapply(proximity_maps, "[[", "BOX")
  ggsave(filename = "figures/prox_maps.pdf", plot = marrangeGrob(proximity_maps.box, nrow = 1, ncol = 1),
         width = 11, height = 8.5)
}

## Spatial Maps, trips by spatial cell ----

# Determine whether the spatial arrangement of monitoring was distributed evenly (as expected by random sampling). 
# Counts the `number of monitored trips in each HEX_ID` and compares to the simulation outcomes, highlighting any actual outcomes that were more extreme than 95% of simulated outcomes. Cells start to get color if they are more extreme than 80% of outcomes.
#' *replaced by plt.spatial_coverage for 2024 AR*
spatial_plots <- plot_monitoring_spatial(box_def, realized_mon, sim.realized.stratum, strata_levels, ak_map.low_res.lst)
plt.spatial <- spatial_plots$plt.spatial.2024
# Drop unused facet
# drop_facet(plt.spatial)
plt.spatial <- drop_facet(plt.spatial, drop = list(c(24, 7), c(25, 7)))

## Coverage maps by spatial cell ----
# These maps use the results of the proximity analyses and summarizes them spatially so they match more closely with the results of the spatiotemporal analyses. Cells that are colored have realized proximity indices in the most extreme 10% of simulation outcomes and those in the most extreme 2.5% are marked with a circle.
plt.spatial_coverage <- plot_spatial_coverage(
  box_def, realized_mon, sim.realized.stratum, sim.programmed.stratum, strata_levels, ak_map.low_res.lst)$coverage
plt.spatial_coverage <- drop_facet(plt.spatial_coverage, drop = list(c(24, 7), c(25, 7)))
```

Under a random selection of trips and deliveries the spatiotemporal distribution of monitoring in a stratum should reflect the spatiotemporal distribution of all trips in the stratum. The evaluation methods here are adapted from the proximity index described in the 2024 Draft ADP (NMFS 2023b). The proximity index was defined as the proportion of sample units in a stratum that were either monitored or near a monitored sample unit in space and time. By considering sample units that were neither monitored nor neighboring a monitored trip as a gap in monitoring, the proximity index quantifies the spatiotemporal extent of monitoring coverage for each stratum, and these results are provided in section 3.4.3. However, the spatiotemporal extent of monitoring can also be summarized spatially within a stratum to indicate whether there were spatial patterns in the distribution of the achieved coverage. The method described below applies to both analyses.\
\
To calculate the proximity index, sample units were placed into spatiotemporal boxes defined by 200-km hexagonal spatial cells (cells) and 1-week time periods. Sample units were allowed to span multiple spatiotemporal boxes and gear types and contribute equally to each box (e.g., a trip that crosses three boxes is counted as 0.33 trips in each box). Sample units were identified as monitored or unmonitored using actual or simulated outcomes. Boxes that either contained a monitored sample unit or were immediately adjacent to a box with a monitored sample unit in both space and time were identified. The number of sample units either inside or adjacent to sample boxes were then totaled. Simulations of random sampling were repeated 10,000 times each using the programmed selection rates and realized monitoring rates. The spatial analyses quantified the proportion of sample units in each spatial cell that were either monitored or adjacent to a monitored sample unit whereas the spatiotemporal analyses calculated this total across the entire stratum (i.e., the proximity index).\
\
Under a random selection of trips and deliveries the spatial distribution of monitored trips should reflect the spatial distribution of all trips. To evaluate whether the actual spatial distribution of monitoring matched what would be expected given the coverage rates that were actually achieved, for each cell, the proportion of sampling iterations that were more extreme than the actual value was calculated to indicate the likelihood of the achieved outcome. By mapping out these deviations, regions where either more or less coverage was achieved than expected are identified.\
\
Spatial biases in the distribution of coverage were not apparent in five of the seven monitored partial selection strata. Most strata had at least a few spatial cells where the fishing effort was either over- or under-represented by the samples, but no clear spatial patterns were apparent to indicate biases. However, spatial biases were apparent in the *OB FIXED BSAI* and the *EM FIXED BSAI* strata. In both strata, the fishing effort in the eastern Aleutians was not as well-represented by the sampled trips as would be expected from random sampling. The spatial distribution of monitoring was as expected in the *OB TRW BSAI* stratum `r fig_nums('plt.spatial_coverage', display = "cite")`. In the *EM FIXED BSAI* stratum, the previously mentioned temporal bias in the video review of this stratum further led to this spatial bias as well.

## Spatiotemporal Distribution of Monitoring Coverage
In the `r year` ADP, the proximity allocation algorithm calculated the expected proximity indices of each stratum as a weighting factor to determine the sample rates necessary to reduce spatiotemporal gaps. To summarize whether the achieved spatiotemporal extent of monitoring coverage in each stratum met expectations, the proximity indices achieved by monitoring were compared to those derived from sampling simulations. The methods for this analysis were described in the previous section, 3.4.2. Again, the proximity index was calculated as the sum of the weight sample units that were either monitored or neighboring monitored sample units divided by the total number of sample units in the stratum. Distributions of proximity indices were generated by simulating sampling at both the selection rates programmed into ODDS as well as the coverage rates actually achieved. The proportions of sampling iterations that were more extreme than the actual values were calculated to indicate the likelihood of the achieved outcome. Comparisons versus simulations using the rates programmed into ODDS inform whether the spatiotemporal extent of monitoring met the expectations planned by the `r year` ADP. Comparisons versus the simulations using the actual coverage rates inform whether there were temporal and/or spatial biases in the distribution of the achieved coverage (`r fig_nums('plt.proximity.stratum', display = "cite")`) for the `r year` ADP (NMFS 2023b).\
\
The spatiotemporal distributions of monitoring coverage met expectations in all strata were within the expected with only one exception ( `r fig_nums('plt.proximity.stratum', display = "cite")`. Six out of seven strata achieved proximity indices of at least 0.92 indicating good spatiotemporal overlap of monitored trips and deliveries with unmonitored trips and deliveries. In other words, in most strata, over 92% of trips were either monitored or within 200 km and 1 week of a monitored trip. However, the *EM FIXED BSAI* stratum only achieved a proximity index of 0.64, much lower than the ranges expected by the selection rates set by the ADP and programmed in ODDS (blue distributions). As mentioned previously, a temporary shortage in reviewer capacity resulted in reduced review for the *EM FIXED* stratum, but it also resulted in a temporal bias. Further analysis also indicated that in this stratum, the lack of reviewed trips disproportionately affected pot gear trips, resulting specifically in data gaps for pot gear EM trips in the BSAI from mid-March to November. The proximity value for the *EM TRW GOA (EFP)* stratum was low but not unexpected, and resulting proximity values were very close to 1 (indicating good spatiotemporal overlap).

## Trip Metrics  
This section analyzes whether monitored trips are similar to unmonitored trips using a permutation test (a.k.a., randomization test). This test evaluates the question “How likely is the difference we found if these two groups have the same distribution (in the metric we are comparing)?” Permutation tests compare the actual difference found between two groups to the distribution of many differences derived by randomizing the labels defining the two groups (e.g., monitored and unmonitored). Difference values in the permutation test were calculated by subtracting the mean metric value for the 'No' condition from the mean metric value for the 'Yes' condition. For example, the difference between vessel lengths in a permutation test for a monitoring effect would be the mean value for unmonitored trips subtracted from the mean value for all monitored trips. By randomizing group assignments, the combined distribution of randomized differences represents the sampling distribution under the null hypothesis that the two groups are equal. In this report, 1,000 randomized trials were run for the permutation test. The *p*-value from the test is calculated as the number of randomized trials with greater absolute differences than the actual difference divided by the number of randomized trials. Similar to the other statistical tests used in this report, low *p*-values indicate unlikely events under the hypothesis of equality and are therefore considered evidence against that hypothesis. However, unlike other statistical tests used in this report, a Bonferroni adjustment has been applied to the significance threshold of 0.05 by dividing it by the number of metrics being tested. This results in an adjusted significance threshold of 0.05 / 6 = 0.00833. The *p*-values are then compared to the adjusted significance threshold. In an attempt to improve clarity, five values are calculated in the test: (1) the difference between groups, (2) the mean difference between groups from randomized trials, (3) #1 expressed as a percentage of the mean value of the metric being tested, (4) #2 expressed as a percentage of the mean value of the metric being tested, and (5) the *p*-value of the test; however, only values (1), (3), and (5) are presented.\
\
Six trip metrics were examined in the permutation test. These metrics were as follows: the number of NMFS Areas visited in a trip, trip duration (days), the weight of the landed catch (t), the vessel length (ft), the number of species in the landed catch, and the proportion (0 to 1) of the total catch that is made up of the most predominant species (pMax). The metric ‘vessel length’ is used to help interpret the results from ‘weight of landed catch’ since fishing power is positively correlated to vessel length. Specifically, differences in weight and length are interpreted as a failure to achieve a random sample of vessels of different sizes, whereas differences in weight only lend more evidence that there was a monitoring effect. The number of species within the landed portion of the catch is a measure of species richness. Our pMax metric follows the concepts behind Hill’s diversity number *N*~1~ that depicts the number of abundant species (Hill 1973) and is a measure of how “pure” catch is since a value of one would indicate that only the predominant (and presumed desirable) species was landed.  

### Were monitored trips similar to unmonitored trips?  

```{r perm_results, set.seed(49), message=FALSE, warning=FALSE, include=FALSE}

dat <- work.data %>% 
  filter(ADP == year & STRATA %in% partial$STRATA)

#' *==================*
#' *2024 ONLY -- Apply the sample unit fix for the EM TRW GOA (EFP) stratum*
# Count of unique TRIP_IDs and observed TRIP_IDs
# dat[STRATA == "EM TRW GOA (EFP)", .(TRIP_ID = uniqueN(TRIP_ID), OBSERVED_Y = uniqueN(TRIP_ID[OBSERVED_FLAG == "Y"]))]
# Update the values using REPORT_ID
dat.em_trw_goa <- dat[STRATA == "EM TRW GOA (EFP)"]
dat.em_trw_goa[, c("TRIP_ID", "OBSERVED_FLAG") := work.data.tem.join[dat.em_trw_goa, .(TRIP_ID, OBSERVED_FLAG), on = .(REPORT_ID)]]
dat <- rbind(dat[STRATA != "EM TRW GOA (EFP)"], dat.em_trw_goa)
# dat[STRATA == "EM TRW GOA (EFP)", .(TRIP_ID = uniqueN(TRIP_ID), OBSERVED_Y = uniqueN(TRIP_ID[OBSERVED_FLAG == "Y"]))]
#' *==================*

# Identify the field to permute by
YN_var = "OBSERVED_FLAG"

# Identify the fields to group by (don't include TRIP_ID or the YN_var)
gp_vec = c("STRATA")

# Run permutation tests
# TODO: ℹ Please use `all_of()` or `any_of()` instead.
  # Was:
#  data %>% select(YN_var)

  # Now:
#  data %>% select(all_of(YN_var))

#See <https://tidyselect.r-lib.org/reference/faq-external-vector.html>.
perm <- permutation.fxn(dat, YN_var, gp_vec, 1000)

# Decompose the output
assign(paste(YN_var, "YN_perm", sep = "."), data.frame(perm[1]))
assign(paste(YN_var, "summary", sep = "."), data.frame(perm[3]))
assign(paste(YN_var, "results", sep = "."), data.frame(perm[4], check.names = FALSE))
assign(paste(YN_var, "Ntable", sep = "."), data.frame(perm[5]))

names(OBSERVED_FLAG.Ntable) <- c("Strata", "Observed", "Unobserved")
OBSERVED_FLAG.Ntable <- arrange(OBSERVED_FLAG.Ntable, match(Strata, strata.order))
OBSERVED_FLAG.summary <- arrange(OBSERVED_FLAG.summary, match(STRATA, strata.order), variable)
perm_sum <- filter(OBSERVED_FLAG.summary, STRATA != "EM TRW EFP")

# See helper.r documentation on interpret_trip_metrics() 
perm_sum <- interpret_trip_metrics(dat = perm_sum)

OBSERVED_FLAG.results <- OBSERVED_FLAG.results %>% mutate(Metric = replace(Metric, Metric == "Observed difference", "OD"))  # Shorten metric
OBSERVED_FLAG.results <- OBSERVED_FLAG.Ntable %>% 
  merge(OBSERVED_FLAG.results, by = "Strata") %>%           # Merge in trip counts of observed/unobserved
  mutate(Observed = formatC(Observed, big.mark = ","),      # Format trip counts
         Unobserved = formatC(Unobserved, big.mark = ","))

# REFORMAT RESULTS TABLE:
OBSERVED_FLAG.results <- rbind(
  OBSERVED_FLAG.results %>% 
    filter(Metric %in% c("OD", "OD (%)")) %>% 
    mutate(across(unique(OBSERVED_FLAG.YN_perm$variable), ~ format(.x, nsmall = 3, digits = 3))),
  OBSERVED_FLAG.results %>% 
    filter(Metric %in% c("p-value")) %>% 
    # Test against Bonferroni-adjusted alpha (0.05/ number of metrics)
    mutate(across(unique(OBSERVED_FLAG.YN_perm$variable), ~ ifelse(as.numeric(.x) < 0.05/length(unique(OBSERVED_FLAG.YN_perm$variable)),
                                                                   paste0(format(.x, nsmall = 3, digits = 3), "*"), 
                                                                   format(.x, nsmall = 3, digits = 3))))) %>% 
  arrange(match(Strata, strata.order), Metric)

# Replace "0" p-values; extra space in character is an artifact of the previous format()

OBSERVED_FLAG.results[OBSERVED_FLAG.results == "0.000*"] <- "< 0.001*"
OBSERVED_FLAG.results[OBSERVED_FLAG.results$Metric == "p-value",]$Metric <- "*p*-value"
```

The sample sizes available and the results of permutation tests are presented in `r tbl_nums('perm.summary.table', display = "cite")`. A visual depiction of individual results of this permutation test is given in `r fig_nums('Obs.perm.plot', display = "cite")` for illustration purposes. Observed trips in the *OB FIXED GOA* stratum were 10.4% (0.6 days) shorter in duration and were on vessels 4.9% (2.8 feet) longer than unmonitored trips. Monitored trips in the EM FIXED BSAI stratum were on vessels 16.2% (14.0 feet) shorter than unmonitored trips. Monitored trips in the EM FIXED GOA stratum landed 24.3% (0.9 species) more species than unmonitored trips. While not directly comparable, the previous hook-and-line observer and EM strata showed similar patterns in 2023, with shorter trips in the observer stratum and more species landed in the EM stratum during monitored trips (AFSC and AKRO 2024).

# Response to Council and SSC Comments

The Council reviewed the `r year` ADP as part of their 11 June 2022 meeting and made the following recommendation:\
\
*The Council supports maintaining the stratification and allocation strategy from the 2022 ADP in 2023. The Council also supports (1) additional fixed gear EM vessels (30% coverage) in the EM pool in 2023 (up to 200 total vessels) provided they opt-in prior to 1 November 2022, additional funding for EM equipment is secured, and they meet the criteria in the ADP and (2) continuation of the pelagic trawl EM project with 100% at-sea monitoring in addition to shoreside observer coverage.*\
\
The SSC has requested that a specific section with responses to SSC comments be provided in the written report, as is done for SAFE documents. The SCC last reviewed the results of this chapter during the presentation of the 2017 Annual Report made at the June 2018 Council meeting.\
\
In January 2022, the SSC reviewed a presentation made by industry and the AKRO on the proposed elements of what is now the *EM TRW EFP* stratum. Items raised related to the replacement of at-sea observers with EM included:

1. The pollock stock assessment team should be closely consulted concerning whether loss of haul level spatial information will impact any ongoing or future analyses and how the data changes will be treated in the assessment (e.g., effective sample sizes for biological collections, weighting of samples from tenders vs individual vessels). It will be essential that authors are prepared to incorporate these new data streams before they become permanent rather than being done in a post hoc manner.\
2. Provide details on how catches and biological data could be assigned to trip- or haul-level information when catches from multiple catcher vessels are mixed on tenders, or how pooled data can be tracked and analyzed appropriately.\
3. Confirmation that this program will not result in a loss of overall specimen and biological samples, particularly in the GOA where this can be most challenging.\
4. Evaluate the potential for large shifts in discard estimates during the year within CAS as compliance monitoring is completed on video review.\
5. Provide more detailed numbers in the next iteration, including examples of biological samples before and after the EFP.\
6. An illustrative example of how salmon PSC calculations would be different under this program would be helpful, including a GOA and BSAI example.

The data used in this and prior Annual Reports are available to answer items 2, 3, 5 and 6 should staff resources at the AKRO and FMA be allocated to this task.

\newpage

# Citations
AFSC (Alaska Fisheries Science Center) and AKRO (Alaska Regional Office). 2024. North Pacific Observer Program 2023 Annual Report. Alaska Fish. Sci. Cent., NOAA, Natl. Mar. Fish. Serv., 7600 Sand Point Way N.E., Seattle, Washington 98115. Available online at: https://www.fisheries.noaa.gov/resource/document/north-pacific-observer-program-2023-annual-report

AFSC (Alaska Fisheries Science Center). 2023. 2024 Observer Sampling Manual. Fisheries Monitoring and Analysis Division, North Pacific Groundfish Observer Program. AFSC, 7600 Sand Point Way N.E., Seattle, Washington, 98115. Available online at: REPLACE LINK https://s3.amazonaws.com/media.fisheries.noaa.gov/2023-06/2023-Observer-Sampling-Manual-508-5-3-23.pdf

Cahalan, J. and C. Faunce. 2020. Development and Implementation of a Fully Randomized Sampling Design for a Fishery Monitoring Program. Fishery Bulletin 118:87–99. doi: 10.7755/FB.118.1.8.

Cahalan, J., J. Mondragon, and J. Gasper. 2014. Catch Sampling and Estimation in the Federal Groundfish Fisheries Off Alaska: 2015 Edition. U.S. Dep. Commer., NOAA Tech. Memo. NMFS-AFSC-286, 46 p.

Faunce, C. H. 2015. Evolution of Observer Methods to Obtain Genetic Material from Chinook Salmon Bycatch in the Alaska Pollock Fishery. U.S. Dep. Commer., NOAA Tech. Memo. NMFS-AFSC-288, 28 p.

`r if(tech.memo){'Faunce, C. H., and Barbeaux, S. J. 2011. The frequency and quantity of Alaskan groundfish catcher-vessel landings made with and without an observer. ICES J. Mar. Sci. 68:1757-1763.'}`  

Hill, M.O. 1973. Diversity and Evenness: A Unifying Notation and Its Consequences. Ecology 61:225–236. doi: 10.2307/1934352.

`r if(tech.memo){'Nelson Jr., R., R. French, R. and J. Wall. 1981. Sampling by U.S. observers on foreign fishing vessels in the eastern Bering Sea and Aleutian Island region, 1977-78. Mar. Fish. Rev. 43:1-19.'}`  

NMFS (National Marine Fisheries Service). 2023a. Draft 2024 Annual Deployment Plan for Observers and Electronic Monitoring in the Partial Coverage Groundfish and Halibut Fisheries off Alaska. National Oceanic and Atmospheric Administration, 709 West 9th Street, Juneau, Alaska 99802. Available online at: https://www.fisheries.noaa.gov/resource/document/draft-2024-annual-deployment-plan-and-partial-coverage-cost-efficiencies-analysis

NMFS (National Marine Fisheries Service). 2023b. 2024 Annual Deployment Plan for Observers and Electronic Monitoring in the Groundfish and Halibut Fisheries off Alaska. National Oceanic and Atmospheric Administration, 709 West 9th Street, Juneau, Alaska 99802. Available online at: https://www.fisheries.noaa.gov/s3//2023-11/Final-2024-ADP.pdf

NMFS (National Marine Fisheries Service). 2022. 2023 Annual Deployment Plan for Observers and Electronic Monitoring in the Groundfish and Halibut Fisheries off Alaska. National Oceanic and Atmospheric Administration, 709 West 9th Street, Juneau, Alaska 99802. Available online at: https://www.fisheries.noaa.gov/resource/document/2023-annual-deployment-plan-observers-and-electronic-monitoring-groundfish-and

NMFS. 2013. 2014 Annual Deployment Plan for Observers in the Groundfish and Halibut Fisheries off Alaska. National Oceanic and Atmospheric Administration, 709 West 9th Street, Juneau, Alaska 99802. Available online at: https://media.fisheries.noaa.gov/dam-migration/final2014adp.pdf

\newpage

`r if(tech.memo){'# Tables'}`

`r tbl_nums('trip.budget.tbl', caption = paste0('Comparison between predicted and actual fishing effort (by number of catcher vessel trips, N) and monitored days at sea (d) for the partial coverage strata in ', year,'.  Predicted values come from Table B-3 of the ', year,' Annual Deployment Plan (ADP).'))`  

```{r trip.budget.tbl, tab.cap=""}

trip.budget.table <- trip.budget.table %>%
  mutate(pred_days = as.numeric(pred_days),
         act_days = as.numeric(act_days),
         pred_N = as.numeric(pred_N),
         act_N = as.numeric(act_N)) 

flextable(trip.budget.table) %>%
  italic(j = 1, i = ~ !(STRATA %like% "total|Total")) %>%
  bold(bold = TRUE, part = "header") %>%
  hline(i = v_diff(trip.budget.table$STRATA, 1, 2)) %>%
  hline(i = which(trip.budget.table[-1, "STRATA"] %like% "total"), border = fp_border(style = "dashed")) %>%
  hline(i = which(trip.budget.table[, "STRATA"] %like% "total")) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  add_header_row(values = c("", "Total trips",   "Monitored/reviewed days"), colwidths = c(1, 4, 4)) %>%
  mk_par(part = "header", i = 1, j = 2, value = as_paragraph("Total trips (", as_i("N"), ")")) %>%
  mk_par(part = "header", i = 1, j = 6, value = as_paragraph("Monitored/reviewed days (", as_i("d"), ")")) %>%
  align(j = 2:9, part = "all", align = "right") %>%
  align(i = 1, part = "header", align = "center") %>%
  width(j = 1, 1.5, unit = "in") %>%
  set_header_labels(values = list(
    STRATA = "Strata",
    pred_N = "Predicted", act_N = "Actual", diff.N = "Difference", perc.N = "Percent",
    pred_days = "Predicted", act_days = "Actual", diff = "Difference", perc = "Percent")) %>%
  width(j = 2, 0.8, unit = "in") %>%
  width(j = 6, 1.1, unit = "in")
```

\endLandscape

`r tbl_nums('disposition.table' , caption = paste0('Trip cancellation rates in the Observer Declare and Deploy System (ODDS) for ', year, '. A trip is canceled by the system if the user did not identify whether fishing had occurred by the end of the year. “Paper” indicates that a trip was logged when ODDS was not available.'))` 
  
```{r disposition.table, tab.cap=""}
# Table 2 from 2019 Annual Report.
# Note that this table (and table 3) exclude any trips that requested coverage (compliance strata)

disposition.table <- disposition.table %>%
  mutate(Year = as.factor(Year),
         `Logged (*a*)` = as.numeric(`Logged (*a*)`),
         `Trips remaining (*c* = *a*-*b*)` = as.numeric(`Trips remaining (*c* = *a*-*b*)`)) %>%
  arrange(factor(Strata, levels = strata.order)) %>%
  filter(Year == year) %>%
  select(-Year)

flextable(disposition.table) %>%
  mk_par(j = 3, part = "header",
         value = as_paragraph("Logged (", as_i("a"), ")")) %>%
  mk_par(j = 4, part = "header",
         value = as_paragraph("Canceled by system (", as_i("b"), ")")) %>%
  mk_par(j = 5, part = "header",
         value = as_paragraph("Trips remaining (", as_i("c"), " = ", as_i("a"), " - ", as_i("b"), ")")) %>%
  mk_par(j = 6, part = "header",
         value = as_paragraph("Canceled by user (", as_i("d"), ")")) %>%
  mk_par(j = 8, part = "header",
         value = as_paragraph("% user cancellation (", as_i("d"), "/", as_i("c"), " \U00D7 100)")) %>%
  italic(j = 1, i = ~Strata != "Total") %>%
  align(j = 3:8, part = "all", align = "right") %>%
  valign(part = "header", valign = "bottom") %>%
  bold(bold = TRUE, part = "header") %>%
  merge_v(j = 1) %>%
  hline(i = seq(3, nrow(disposition.table), 3)) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  fix_border_issues() %>%
  width(j = 1, 1.23, unit = "in") %>%
  width(j = 2, 1.38, unit = "in") %>%
  width(j = 3, 0.88, unit = "in") %>%
  width(j = 4, 1, unit = "in") %>%
  width(j = c(5:6), 0.94, unit = "in") %>%
  width(j = 7, 0.56, unit = "in") %>%
  width(j = 8, 0.63, unit = "in")
```

\newpage

`r tbl_nums('logged.table', caption = paste0('Number of remaining trips after cancellation in each trip-selection stratum that were selected using the initial random number generator ("Random number selection") and those that remained after user manipulation ("Total final selected") in ', year, '. The relative impact of waivers in trip-selection is also shown ("% reduction of selected trips due to waivers).'))` 

```{r logged.table, tab.cap=""}
# Table 3 from 2019 Annual Report
# Note that this table (and table 2) exclude any trips that requested coverage (compliance strata)

logged.table <- logged.table %>%
  mutate(`Total completed trips` = as.numeric(`Total completed trips`)) %>%
  arrange(factor(Strata, levels = strata.order)) %>%
  ungroup()

flextable(logged.table) %>%
  mk_par(j = 3, part = "header", value = as_paragraph("Selection type")) %>%
  mk_par(j = 4, part = "header", value = as_paragraph("Selected trips")) %>%
  mk_par(j = 5, part = "header", value = as_paragraph("Waived trips")) %>%
  mk_par(j = 6, part = "header", value = as_paragraph("Total final selected")) %>%
  mk_par(j = 7, part = "header", value = as_paragraph("% selected from inherits")) %>%
  mk_par(j = 8, part = "header", value = as_paragraph("% reduction of\n selected trips\n due to waivers")) %>%
  align(j = ~ . - Strata, part = "all", align = "right") %>%
  valign(part = "header", valign = "bottom") %>%
  italic(j = 1, i = 1:6) %>%
  bold(bold = TRUE, part = "header") %>%
  #hline(i = seq(from = 2, to = nrow(logged.table), by = 2), border = fp_border(style = "dashed")) %>%
  hline(i = v_diff(logged.table$Strata)) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  # Merge repeated values, grouped by Strata for the columns listed in 'target'
  merge_v(j = "Strata", target = c("Strata", "Total completed trips", "Total final selected", 
                                   "% Selected from inherits", "% Reduction of selected trips due to waivers")) %>%
  fix_border_issues() %>%
  width(j = 1, 1.3, unit = "in") %>%
  width(j = c(2, 5), 0.94, unit = "in") %>%
  width(j = 3, 1.31, unit = "in") %>%
  width(j = c(4, 6:7), 1.06, unit = "in") %>%
  width(j = 8, 1.2, unit = "in")
```

\newpage

`r tbl_nums('Selection.table', caption = paste0('Number of logged trips in each partial coverage stratum in ', year, ' that were selected using the initial random number generator ("Initial random selection") and those that remained after user manipulation ("After cancellations"). The relative impact of inherits and waivers in trip-selection is also shown ("With inherits", "After waivers").'))`  

```{r Selection.table, tab.cap=""}

flextable(Selection.table) %>%
  set_header_labels(values = list(
    n = "Selected trips",
    N = "Total trips",
    Actu = "Actual (%)",
    Prog = "Programmed (%)")) %>%
  mk_par(j = 7, part = "header",
         value = as_paragraph(as_i("p"), "-value")) %>%
  mk_par(j = 2, i = seq(1, nrow(Selection.table), 4), part = "body",
         value = as_paragraph("Initial random selection, ", as_i("a"))) %>%
  mk_par(j = 2, i = seq(2, nrow(Selection.table), 4), part = "body",
         value = as_paragraph("After cancellations, ", as_i("b"), " (", as_i("a"), " - ", as_i("b"), ")")) %>%
  mk_par(j = 2, i = seq(3, nrow(Selection.table), 4), part = "body",
         value = as_paragraph("With inherits, ", as_i("c"), " (", as_i("a"), " - ", as_i("b"), " + ", as_i("c"), ")")) %>%
  mk_par(j = 2, i = seq(4, nrow(Selection.table), 4), part = "body",
         value = as_paragraph("After waivers, ", as_i("d"), " (", as_i("a"), " - ", as_i("b"), " + ", 
                              as_i("c"), " - ", as_i("d"), ")")) %>%
  align(j = 3:7, part = "all", align = "right") %>%
  valign(part = "header", valign = "bottom") %>%
  align(j = 2, align = "left") %>%
  bold(bold = TRUE, part = "header") %>%
  italic(j = 1) %>%
  merge_v(j = 1) %>%
  hline(i = seq(4, nrow(Selection.table), 4)) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  fix_border_issues() %>%
  width(j = 1, 1.23, unit = "in") %>%
  width(j = 2, 2.06, unit = "in") %>%
  width(j = 3, 1.06, unit = "in") %>%
  width(j = 4:5, 0.88, unit = "in") %>%
  width(j = 6, 1.31, unit = "in") %>%
  width(j = 7, 0.69, unit = "in")
```

\newpage

`r tbl_nums('Coverage.Table', caption = paste0('Number of total vessels (*V*), monitored vessels (*v*), total trips/deliveries (*N*) and monitored trips/deliveries (*n*) for each stratum in ', year, '. The coverage and 95% confidence interval columns are expressed as percentages of the total number of trips taken/deliveries made within each stratum. \\* indicates where *N* and *n* represent deliveries'))`  

```{r Coverage.Table, tab.cap=""}

Coverage.Table <- Coverage.Table %>% 
  mutate(Strata = as.character(Strata),
         V = as.numeric(V),
         v = as.numeric(v),
         N = as.numeric(N),
         n = as.numeric(n)) %>% 
  mutate(Strata = ifelse(Coverage %in% c("Full coverage total", "Partial coverage total"), as.character(Coverage), Strata)) %>%
  mutate(Coverage = case_when(Coverage == "Partial coverage total" ~ "Partial coverage",
                              Coverage == "Full coverage total" ~ "Full coverage",
                              Coverage == "Total" ~ "Zero coverage",
                              TRUE ~ Coverage)) %>%
  arrange(factor(Strata, levels = strata.order)) %>%
  arrange(factor(Coverage, levels = c("Full coverage", "Partial coverage", "Zero coverage")))

as_grouped_data(Coverage.Table, groups = "Coverage") %>%
  as_flextable(hide_grouplabel = TRUE) %>%
  bold(bold = TRUE, part = "header") %>%
  bold(i = ~ !is.na(Coverage)) %>%
  bold(j = 1, i = ~ Strata == "Total") %>%
  italic(j = 1, i = ~ !is.na(Strata) & !(Strata %like% "total") & Strata != "Total") %>%
  italic(j = c("V", "v", "N", "n"), part = "header") %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  flextable::surround(i = ~ !is.na(Coverage) | Strata == "Total",
                      border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  flextable::surround(i = ~ Strata %like% "total", border.top = fp_border(width = 1, style = "dashed")) %>%
  add_header_row(values = c("", "Coverage", "95% confidence interval", ""), colwidths = c(5, 2, 2, 1)) %>%
  width(j = 1, 1.6, unit = "in") %>%
  width(j = 2:3, 0.38, unit = "in") %>%
  width(j = 4:5, 0.5, unit = "in") %>%
  width(j = 6:7, 0.75, unit = "in") %>%
  width(j = 8:9, 0.58, unit = "in") %>%
  width(j = 10, 1.77, unit = "in") %>%
  align(j = 2:10, part = "all", align = "right") %>%
  align(i = 1, part = "header", align = "center") %>%
  # NOTE: this breaks if you try and run it here, but knits correctly
  #  - These 3 lines ensure the long text in the "Total" row aligns correctly
  #' \TODO *TEST THIS ONCE 1_AR IS RE-RUN AND WE CAN KNIT AND TEST THIS*
  merge_h_range(i = 16, j1 = 7, j2 = 10) %>%
  align(i = 16, j = 7, align = "left", part = "body") %>%
  padding(i = 16, j = 7, padding.left = 25, part = "body") %>%
  valign(part = "header", valign = "bottom")
```

\newpage

`r tbl_nums('Pollock.observed.table', caption = paste0('The number of shoreside pollock deliveries made by catcher vessels in the *OB TRW* strata during ', year, ', separated by port and coverage category. Trips that made a delivery to a tender have been excluded. Observed deliveries denote deliveries that were monitored shoreside for salmon.'))`  

```{r Pollock.observed.table, tab.cap=""}

Pollock.observed.table <- 
  Pollock.observed.table %>%
  mutate(`Total deliveries (*N*)` = as.numeric(`Total deliveries (*N*)`),
         `Observed deliveries (*n*)` = as.numeric(`Observed deliveries (*n*)`))

flextable(Pollock.observed.table) %>%
  mk_par(j = 4, part = "header",
         value = as_paragraph("Total deliveries (", as_i("N"), ")")) %>%
  mk_par(j = 5, part = "header",
         value = as_paragraph("Observed deliveries (", as_i("n"), ")")) %>%
  align(j = 4:6, part = "all", align = "right") %>%
  align(j = 1:3, part = "all", align = "left") %>%
  valign(part = "header", valign = "bottom") %>%
  bold(bold = TRUE, part = "header") %>%
  merge_v(j = c("FMP", "Coverage category")) %>%
  merge_h_range(i = ~ FMP %like% "total", j1 = 1, j2 = 3) %>%
  flextable::surround(i = ~FMP %like% "total", border.bottom = fp_border(width = 1), border.top = fp_border(style = "dashed")) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  fix_border_issues() %>%
  width(j = 1, 0.81, unit = "in") %>%
  width(j = 2, 1.39, unit = "in") %>%
  width(j = 3, 1, unit = "in") %>%
  width(j = 4, 1.44, unit = "in") %>%
  width(j = 5, 1.69, unit = "in") %>%
  width(j = 6, 0.94, unit = "in")
```

\newpage

`r tbl_nums('Pollock.efp.table', caption = paste0('The number of pollock deliveries made by catcher vessels in the *EM TRW (EFP)* strata during ', year, ', separated by coverage category and port. Observed deliveries denote deliveries that were monitored shoreside for salmon. Tender and non-tender denote whether or not a catcher vessel delivered its catch to a tender.'))`  

```{r Pollock.efp.table, tab.cap=""}

Pollock.efp.table <-
  Pollock.efp.table %>%
    mutate(`Total deliveries (*N*)` = as.numeric(`Total deliveries (*N*)`),
         `Observed deliveries (*n*)` = as.numeric(`Observed deliveries (*n*)`))

flextable(Pollock.efp.table) %>%
  mk_par(j = 4, part = "header",
         value = as_paragraph("Total deliveries (", as_i("N"), ")")) %>%
  mk_par(j = 5, part = "header",
         value = as_paragraph("Observed deliveries (", as_i("n"), ")")) %>%
  align(j = 4:6, part = "all", align = "right") %>%
  bold(bold = TRUE, part = "header") %>%
  merge_v(j = c("FMP", "Coverage category")) %>%
  merge_h_range(i = ~ FMP %like% "total", j1 = 1, j2 = 3) %>%
  flextable::surround(i = ~ FMP %like% "total", border.bottom = fp_border(width = 1), border.top = fp_border(style = "dashed")) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
  fix_border_issues() %>%
  width(j = 1, 1.00, unit = "in") %>%
  width(j = 2, 1.39, unit = "in") %>%
  width(j = 3, 1.00, unit = "in") %>%
  width(j = 4, 1.44, unit = "in") %>%
  width(j = 5, 1.69, unit = "in") %>%
  width(j = 6, 0.94, unit = "in") 
  
```

\newpage

`r tbl_nums('perm.summary.table', caption = paste0('Results of permutation tests between monitored and unmonitored trips in the ', year, ' trip-selection strata. OD: Observed difference (monitored - unmonitored). Observed and unobserved columns are in units of trips (or shoreside deliveries for the *EM TRW GOA [EFP]* stratum). Statistically significant results are in bold italics. The Bonferroni correction was applied to the significance tests.'))`  

```{r perm.summary.table, tab.cap=""}

# Identify significant outcomes
OBSERVED_FLAG.results.sig <- lapply(OBSERVED_FLAG.results, function(x) grepl("[*]$", x))
# Apply to raw difference and percentage rows
OBSERVED_FLAG.results.sig <- lapply(OBSERVED_FLAG.results.sig, function(x) x | c(x[-1], F) | c(x[-c(1,2)], F, F))

flextable(OBSERVED_FLAG.results) %>%
  bold(bold = TRUE, part = "header") %>%
  italic(j = 1, part = "body") %>%
  hline(i = seq(3, nrow(OBSERVED_FLAG.results), 3)) %>%
  flextable::surround(part = "header", border.top = fp_border(width = 1), border.bottom = fp_border(width = 1)) %>%
   mk_par(j = 4, i = seq(3, nrow(OBSERVED_FLAG.results), 3), part = "body",
         value = as_paragraph(as_i("p"), "-value")) %>%
  merge_v(j = c("Strata", "Observed", "Unobserved")) %>%
  align(j = 2:10, part = "all", align = "right") %>%
  valign(j = 1:3, part = "body", valign = "top") %>%
  valign(part = "header", valign = "bottom") %>%
  fix_border_issues() %>%
  width(j = 1, 1.6, unit = "in") %>%
  width(j = 2, 0.81, unit = "in") %>%
  width(j = c(3, 6), 0.94, unit = "in") %>%
  width(j = 4, 0.69, unit = "in") %>%
  width(j = 5, 1, unit = "in") %>%
  width(j = c(7, 10), 1.25, unit = "in") %>%
  width(j = 8, 1.13, unit = "in") %>%
  width(j = 9, 1.03, unit = "in") %>%
  bold(j = 5, i = OBSERVED_FLAG.results.sig[[5]]) %>%
  bold(j = 6, i = OBSERVED_FLAG.results.sig[[6]]) %>%
  bold(j = 7, i = OBSERVED_FLAG.results.sig[[7]])%>%
  bold(j = 8, i = OBSERVED_FLAG.results.sig[[8]])%>%
  bold(j = 9, i = OBSERVED_FLAG.results.sig[[9]])%>%
  bold(j = 10, i = OBSERVED_FLAG.results.sig[[10]])
```

\endPortrait

`r if(tech.memo){'# Figures'}`

```{r outcomes.costs.plot, warning=FALSE, fig.width=7.5, fig.height=9, dpi = 600}

# Make a list of grobs of your figures, set the widths as the maximum among the figures.
arrange.plot.fxn <- function(plotA, plotB){
  plot_grobs <- lapply(list(plotA, plotB), ggplotGrob)
  
  lst <- vector(mode = "list", length = 2)
  for(i in seq_along(plot_grobs)) {
  lst[[i]] <- plot_grobs[[i]]$widths
  }
  
  for(i in 1:length(plot_grobs)) plot_grobs[[i]]$widths <- do.call(grid::unit.pmax, lst)
  grid.arrange(plot_grobs[[1]], plot_grobs[[2]], nrow = 2)
}

arrange.plot.fxn(outcomes.plot, costs.plot)
```

`r fig_nums('outcomes.costs.plot', caption = paste0('Total number of observer sea days purchased (top panel) and total cost of observing those sea days (bottom panel). Vertical bars signify the range of potential outcomes predicted by the ', year, ' Annual Deployment Plan. Dashed lines signify expected outcomes. Solid lines signify what actually occurred in ', year, '.'))`  

\newpage

```{r odds.by.date.plot, fig.width=7.5, fig.height=8, dpi=600}
suppressWarnings(print(odds.by.date.plot))
```

`r fig_nums('odds.by.date.plot', caption = paste0('Rate of selected trips logged into the Observer Declare and Deploy System (ODDS) during ', year, ' organized by original date entered for all trips (gray line and gray text), and final date considering only non-canceled trips (black line and black text). The programmed selection rate is depicted as the dotted line. Gray shaded areas denote the range of coverage rate corresponding to the 95% confidence intervals expected from the binomial distribution. Vertical tick marks on the x-axis depict dates when selected trips were canceled (gray, on the bottom) and when inherited trips were monitored (black, on the top).'))` 

\newpage

```{r times_figure, warning=FALSE, fig.width=7.5, fig.height=6, dpi=600, eval=TRUE}
#' \TODO MOVE THIS FIGURE - IT should be second to last, after permutation tests?
suppressMessages(print(times_figure))
```

`r fig_nums('times_figure', caption = "Distributions of data timeliness (the time between a trip or delivery ending and those monitoring data being available for catch accounting) by stratum. Solid black lines and annotations to the left show median data timeliness. Dashed red lines and annotations to the right show mean data timeliness.")`  

\newpage

```{r review_times_figure, warning=FALSE, fig.width=7.5, fig.height=6, dpi=600, eval=TRUE}
#' \TODO MOVE THIS FIGURE - IT should be last, after data timeliness?
suppressMessages(print(review_times_figure))
```

`r fig_nums('review_times_figure', caption = "Distributions of review timeliness (the time between a trip or delivery ending and those trips being reviewed by stratum). Solid black lines and annotations to the left show median review timeliness. Dashed red lines and annotations to the right show mean review timeliness.")`  

\newpage

```{r Temporal.plot, warning=FALSE, fig.width=7.5, fig.height=8, dpi=600}
print(Temporal.plot)
```

`r fig_nums('Temporal.plot', caption = paste0('Cumulative number of trips monitored during ', year, ' (black line) compared to the expected range of observed trips (shaded ribbon) given fishing effort and sampling rates. Dates where the monitored number of trips is outside of expected (less or more than the range) are depicted as tick marks on the x-axis. Test results (using a binomial distribution) determining if the observed rate was sampled at the selection rate are denoted as *p*-values.'))`  

\newpage

```{r plt.spatial_coverage, warning=FALSE, fig.width=7.5, fig.height=8, dpi=600}
print(plt.spatial_coverage)
```

`r fig_nums('plt.spatial_coverage', caption = paste0("Spatial patterns of the distribution of monitoring in partial coverage strata in ", year, " relative to the distribution created from 10,000 simulations of random sampling at each stratum's realized monitoring rate. Each hexagonal spatial cell is 200 km wide. The degree of monitoring in each cell was quantified as the proportion of trips that were monitored or neighboring a monitored trip in both space and time. Cells where the degree of monitoring was more extreme than 90% of simulated outcomes are filled violet (less) or green (more), and those cells with a more extreme outcome than 97.5% of simulated outcomes are additionally marked with a circle."))`

\newpage

```{r plt.proximity.stratum, warning=FALSE, fig.width=7.5, fig.height=8, dpi=600}
print(plt.proximity.stratum)
```

`r fig_nums('plt.proximity.stratum', caption = paste0('Stratum-level proximity indices in partial coverage strata in ', year, '. The purple vertical dashed line represents actual proximity indices. The distributions show the proximity values obtained from 10,000 simulations of random sampling, where the upper (green) distribution sampled using the realized monitoring rate and the lower (blue) distribution used the programmed monitoring rate. The 2.5% tails of the distributions are shaded darker to represent unlikely outcomes. The number of sample units in each stratum is displayed in the upper-left of each facet. Note the varying scales of the x-axes between facets.'))`

\newpage

```{r Obs.perm.plot, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, dpi=600}
# Plot results
OBSERVED_FLAG.YN_perm_plot <- ungroup(OBSERVED_FLAG.YN_perm) %>% mutate(STRATA = as.character(STRATA)) 
OBSERVED_FLAG.summary_plot <- ungroup(OBSERVED_FLAG.summary) %>% mutate(STRATA = as.character(STRATA)) %>% 
  # Apply Bonferroni adjustment to alpha value and format significant p-values  
  mutate(p = ifelse(p < (0.05 / length(unique(OBSERVED_FLAG.summary$variable))), paste0(formatC(p,digits = 3, format = "f"), "*"), formatC(p, digits = 3, format = "f"))) %>%
  mutate(p = ifelse(p == "0.000*", 
                    paste0("italic(p) < '0.001*'"), 
                    ifelse(p < (0.05 / length(unique(OBSERVED_FLAG.summary$variable))), paste0("italic(p) == '", p, "*'"), paste0("italic(p) == '", p, "'"))))

# Filter out EM TRW EFP stratum
# OBSERVED_FLAG.YN_perm_plot <- OBSERVED_FLAG.YN_perm_plot %>% filter(STRATA != "EM TRW EFP")
# OBSERVED_FLAG.summary_plot <- OBSERVED_FLAG.summary_plot %>% filter(STRATA != "EM TRW EFP")

OBSERVED_FLAG.YN_perm_plot$STRATA <- factor(OBSERVED_FLAG.YN_perm_plot$STRATA, levels = strata.order)
OBSERVED_FLAG.summary_plot$STRATA <- factor(OBSERVED_FLAG.summary_plot$STRATA, levels = strata.order)

# Generate permutation plot 
Obs.perm.plot <- ggplot(OBSERVED_FLAG.YN_perm_plot, aes(x = perm_result_pct)) +
  geom_histogram(aes(y = after_stat(density)), col = "white", fill = "gray", bins = 30) +
  geom_density(fill = "gray", alpha = 0.5, adjust = 3) +
  geom_vline(aes(xintercept = 0), col = "black", lty = 3, linewidth = 1) +
  geom_vline(aes(xintercept = obs.diff_pct), col = "red") +
  facet_nested_wrap(. ~ STRATA + variable, scales = "free", 
             # Make number of cols in facet_wrap = to number of strata
             nrow = length(unique(OBSERVED_FLAG.YN_perm_plot$STRATA))) +
  geom_label(data = OBSERVED_FLAG.summary_plot, aes(label = paste(p), fill = factor(p %like% "\\*", levels = c(TRUE, FALSE))), 
             x = -Inf, y = Inf, hjust = -0.1, vjust = 1.2, size = 3, parse = TRUE, label.padding = unit(0.15, "lines")) +
  scale_fill_manual(values = c("pink", "white"), guide = "none") + 
  fig.theme +
  xlab("\nDifference (%) of observed minus unobserved trips relative to the mean") +
  ylab("Density of outcomes\n") + 
  theme(strip.text.x = element_text(size = 8, margin = margin(0.1, 0.1, 0.1, 0.1, unit = "lines"))) + 
  theme(axis.text.y = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

Obs.perm.plot
```

`r fig_nums('Obs.perm.plot', caption = paste0('Results from permutation tests depicting percent differences between monitored and unmonitored trips by strata in the partial coverage category. Gray bars depict the distribution of differences between monitored and unmonitored trips when the assignment of monitoring status has been randomized (this represents the sampling distribution under the null hypothesis that monitored and unmonitored trips are the same). The vertical red solid line denotes the actual difference between monitored and unmonitored trips. Values on the x-axis have been scaled to reflect the relative (%) differences in each metric. The *p*-value for each test is denoted in the upper left corner. Low *p*-values (shaded pink) are reason to reject the null hypothesis and conclude that there is an observer effect.'))`  

\newpage

This page intentionally blank

\newpage

```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
save.image(file = "4_AR_report.Rdata")
```

```{r, compliance check, eval=FALSE}
#' Check of full coverage vessels that failed to take an observer to provide to OLE 
#' Alex Perry - Compliance Analyst - Anchorage [Alex.Perry@noaa.gov]

not_observed <- work.data %>%
  filter(STRATA == "FULL" & ADP == year & OBSERVED_FLAG == "N") %>%
  select(TRIP_ID, VESSEL_ID, PERMIT, LANDING_DATE) %>%
  unique()
```